Enteropathogenic E. coli-mediated Fast and Coordinated Ca²⁺ responses regulate NF-κB activation

Version 1

Abstract

Enteropathogenic Escherichia coli (EPEC) is a major bacterial enteropathogen causing diarrhea among children in developing countries. Here, we found that EPEC induced isolated Ca2+ responses in epithelial cells, triggered by extracellular ATP (eATP). These responses were dependent on type III secretion (T3S) and down-regulated by the bacterial secreted protease EspC, consistent with eATP released by the T3S translocon. By performing high speed Ca2+ imaging, we uncovered that at the onset of infection, low eATP levels triggered Ca2+-responses involving the whole cell but showing the small amplitude and fast kinetics usually associated with local Ca2+ responses. The findings, supported by theoretical modeling, evocate a conceptual shift whereby low amounts of inositol 1, 4, 5-trisphosphate (IP3) induced by low eATP levels and subsequent moderate Ca2+ release enable the fast coordination of IP3 receptor cluster activation throughout the cell. Importantly, these yet undescribed coordinated fast responses occurred over prolonged time periods and defined a cell state with dampened activation of the pro-inflammatory transcriptional activator NF-kB associated with a decrease in its Ca2+-dependent O-linked β-N-acetylglucosamine modification.

Introduction

EPEC are diarrheagenic E. coli strains that cause significant morbidity and mortality in children under two years of age. While infection rates have significantly declined in industrialized nations, EPEC remains a major public health concern in low-income countries (Lozer et al., 2013). EPEC form attaching and effacing (A/E) lesions on intestinal epithelial cells and lack the ability to produce Shiga toxins or heat-labile (LT) and heat-stable (ST) enterotoxins (Croxen et al., 2013; Gomes et al., 2016; Hazen et al., 2016). The ability of EPEC to form A/E lesions is determined by the locus of Enterocyte Effacement (LEE), a large genomic pathogenicity island that encodes the essential genetic elements required for this process (Pearson et al., 2016). The LEE region of EPEC (E2348/69) encodes components of the type III secretion system (T3SS), a molecular apparatus that translocate at least 25 bacterial effector proteins into the host cell. The EPEC type III secretion system (T3SS) consists of a basal body and a needle-like structure, resembling those found in Salmonella and Shigella, but with a distinct sheath-like extension at the needle tip, primarily composed of EspA, and about ten times longer than the T3SS needles from other bacterial species. This EspA filament acts as a molecular bridge, extending from the bacterium to the host cell membrane, allowing insertion of the EspB and EspD translocon components into the host cell membrane, enabling type III effectors injection in the cell cytosol (Creasey et al., 2003; Monjarás Feria et al., 2012; Sal-Man et al., 2012). Osmoprotection assays suggest that the translocon forms a pore with an internal diameter ranging between 3 to 5 nm, allowing the passage of unfolded effector proteins into the host cell (Chatterjee et al., 2015). However, the T3SS translocon shows weak pore-forming activity during EPEC infection of epithelial cells, presumably because it forms a sealed conduct between the T3SS and host cell membranes (Guignot et al., 2016). EspC, a secreted serine protease from the autotransporter family, targets EspA and EspD and down-regulates pore formation activity associated with cytotoxicity (Guignot et al., 2015). EPEC T3SS effector proteins translocated into host cells lead are responsible for attaching and effacing (A/E) lesions and intimate bacterial adhesion to the host cells associated with the formation of an actin-rich pedestal formation (Chen & Frankel, 2005).

The detection of pathogenic bacteria by intestinal epithelial cells plays an important role in initiating pro-inflammatory responses. Recognition of bacterial surface components by pattern recognition receptors triggers pro-inflammatory signaling pathways involving the transcriptional activator NF-κB and the production of cytokines such as interleukin-8 and tumor necrosis factor-α (TNF-α) (Edwards et al., 2011). However, EPEC suppresses these signaling pathways early in infection through the coordinated action of several T3SS effector proteins. Among the first translocated T3SS effectors, Tir interacts with TNF-α receptor-associated factors (TRAF2 and TRAF6), recruiting the tyrosine phosphatases SHP-1 and SHP-2 (Mills et al., 2008; Ruchaud-Sparagano et al., 2011; Yan et al., 2013). Several non-LEE T3SS effectors, including NleE, NleB, NleH1, NleH2, NleC, and NleD, further contribute to inhibiting NF-κB and MAPK signaling. NleE and NleB stabilize the interaction between NFκB and its inhibitory subunit IκB, preventing its degradation, thereby keeping NFκB in an inactive state. The ability of NleE to inhibit NFκB signaling depends on its S-adenosyl-L-methionine (SAM)-dependent methyltransferase activity (Zhang et al., 2011). NleE-mediated methylation of TAB2/3 prevents IKK activation (Zhang et al., 2011). NleB selectively blocks NFκB activation through GlcNAcylation of the TNF-α receptor (TNFR) adaptor protein and TNFR1-associated death domain (TRADD) (S. Li et al., 2013; Pearson et al., 2013). The T3SS effectors NleH1, NleH2 and NleC also interfere with NFκB nuclear translocation (Gao et al., 2009). NleC specifically cleaves P65 RelA (Giogha et al., 2015; Ruchaud-Sparagano et al., 2011; Yen et al., 2010). Interestingly, NleF has been implicated in the activation of NF-κB, underscoring the complexity of the regulation of inflammation during bacterial infection and suggesting the timing of various T3SS effectors’ activity (Pallett et al., 2014).

A similar complexity applies to T3SS effectors regulating cell death and survival pathways during EPEC infection of epithelial cells. Tir was found to elicit a rapid Ca²⁺ influx across the host cell membrane, through the activation of a host plasma membrane Ca²⁺ channel, the mechanosensitive transient receptor potential vanilloid 2 (TRPV2) leading to pyroptosis (Zhong et al., 2022). However, the NleA effector, blocks the delivery of TRPV2 channels to the cell surface, thereby dampening Tir-induced Ca²⁺ influx. The effector NleF also directly binds caspase-4 to inhibit its activity (Zhong et al., 2020). The extrinsic apoptotic pathway is triggered by EPEC pili (Abul-Milh et al., 2001). However, the NleD and NleB effectors inhibit this pathway by cleaving JNK and GlcNAcylating the death domain adaptor proteins TRADD and FADD, respectively (Baruch, Gur-Arie, et al., 2011; Pearson et al., 2013). While EspC prevents cytotoxicity linked to pore-formation by the T3SS translocon during the early EPEC infection phases, it was shown to promote intrinsic apoptosis through increase in intracellular Ca2+ and calpain activation (Serapio-Palacios & Navarro-Garcia, 2016).

Central to inflammation and cell death / survival pathways induced by EPEC, bacterial-induced Ca2+ signals have been a matter of debate. EPEC infection is known to perturb host Ca²⁺ signaling, but the source and sequence of Ca²⁺ signals during infection remain controversial. EPEC was shown to induce Ca²⁺ influx associated with a loss of mitochondrial membranes permeability leading to cell death (Zhong et al., 2020; Ramachandran et al., 2020; Zhong et al., 2022), but was also reported to trigger IP3-mediated Ca²⁺ release possibly involved in bacterial-induced cytoskeletal rearrangements (Baldwin et al., 1991; Baldwin et al., 1993; Foubister et al., 1994; Bain et al., 1998).

Here, we investigated the characteristics and implications of EPEC-induced Ca²⁺ responses in epithelial cells. We characterized yet undescribed Ca²⁺ signals induced by EPEC and low ATP levels, presenting the fast dynamics and small amplitude of local Ca²⁺ responses but involving large cell area. We found that these responses likely result from the coordination of elementary responses via rapid Ca2+-induced Ca²⁺ release over large cell area, challenging generally admitted concepts on Ca²⁺ diffusion. Importantly, we show that these newly described responses have functional implications by dampening the cell ability to respond to inflammatory signals.

Results

EPEC induces Ca²⁺ responses that depend on Type III secretion-mediated eATP release

Despite their critical role in cellular processes key to bacterial infection, EPEC-induced Ca²⁺ responses remain to be characterized. We therefore set up to perform a detailed single cell imaging of Ca²⁺ responses elicited by cells infected by EPEC.

As shown in Fig. 1, EPEC induced Ca²⁺ transients often corresponding to a single peak of varying amplitude detected over several minutes, with a median corresponding to 6.2 ± 0.8% of the maximal histamine response (Figs. 1A, B). These Ca²⁺ responses were dependent on a functional T3SS, since they were not observed for the T3SS-deficient escN mutant (Figs. 1A, C). Only 40 ± 4.7 % of cells, however, elicited responses when challenged wild-type EPEC at a low multiplicity of infection (MOI) of 20 bacteria per cell, a value that raised to 76 ± 4.8 % when using a high MOI of 80 bacteria per cell (Fig. 1C). In contrast, even at the low MOI, more than 83 % of cells showed actin pedestals, indicating that Ca²⁺ responses were elicited only in a fraction of cells targeted by EPEC-type T3SS (Figs. 1D, E). The frequency of Ca²⁺ responses per cell increased over the incubation time with an average frequency of responses per cell raising by 6.9 to 8.3-fold from the first to the last 30 min of EPEC challenge (Fig. 1F). This increased frequency suggested the accumulation of an agonist in the extracellular medium during the course of the infection triggering IP3-mediated Ca²⁺ release. When pooling all single cell responses, we could observe a steady increase in the average cytosolic Ca2+ concentration of the cell population, as previously reported (Ramachandran et al., 2020; Figs. S1B). However, when performing single cell imaging, we did not observe an increase in cytosolic Ca²⁺ basal levels even at high MOI after 2 hours incubation with wild-type bacteria (Fig. S1A).

EPEC induces isolated Ca2+ responses of limited amplitude in epithelial cells

HeLa cells were loaded with the fluorescent indicator Cal-520, challenged with the indicated bacteria and subjected to live-cell Ca2+ imaging at a frequency of one acquisition every 10 seconds (A-C, F) or fixed and processed for fluorescence microscopy analysis (D-E) (Materials and Methods). A, Representative traces of Ca2+ variations in single cells. The black arrowheads indicate the time of bacterial challenge. The blue arrowheads indicate stimulation with 3 µM histamine. B, Response amplitude expressed as a percent of the maximal histamine response amplitude (N = 3, n > 63). C, Percent of cells exhibiting Ca2+ responses (N =3, n > 66). (D, E) Cells challenged with RFP-expressing bacteria for 1 hour. D, Representative confocal micrographs. Staining with DAPI (blue), phalloidin-Alexa 488 (green). The lower panels show a higher magnification of the insets in the top panels. Scale bar = 10 µm. E, Percentage of bacteria-associated actin-rich pedestals (N = 3, n > 273). F, average number of responses per cell during the first 30 min (0-30) and last 30 min (30-60) of bacterial challenge. Low MOI: 10 bacteria / cell. High MOI: 50 bacteria / cell. Bar: mean. (N = 3, n > 63). Mann-Whitney test. *: p < 0.05; **: p < 0.01; ***: p < 0.001; ****: p < 0.0001.

Together, these results suggest that EPEC induces isolated Ca²⁺ responses that depend on the T3SS for the release of limiting amounts of a Ca²⁺ agonist.

EPEC-mediated Ca²⁺ responses depend on ATP released in the extracellular medium via the T3SS translocon

In previous works, we showed that the EPEC T3SS translocon forms pores in host cell plasma membranes that were down-regulated by the bacterial secreted serine protease EspC (Guignot et al., 2016). We posit that low amounts of ATP released in the extracellular medium by the T3SS translocon were responsible for the isolated Ca²⁺ responses of reduced amplitude elicited by EPEC. According to this view, by removing T3SS translocons from host cell membranes, EspC would down-regulate EPEC-mediated Ca2+ signaling explaining the low ratio of Ca2+ responding cells relative to cells forming actin pedestals.

Consistent with this and as shown in Figs. 2A-C, an espC mutant induced more Ca²⁺ responses than wild-type EPEC, with 94 ± 3% responding cells and a frequency of 10.5 ± 1.2 responses per cell over the 60 min analysis, compared 40 ± 4.7% responding cells and less than 2 responses per cell for the espC mutant and wild-type EPEC, respectively. Also, the average amplitude of Ca²⁺ responses induced by the espC mutant was higher than that of wild-type EPEC, suggesting more eATP release (Fig. 1B). Accordingly, cell treatment with Suramin, an inhibitor of purinergic receptors, abolished Ca²⁺ responses induced by the wild-type and espC mutant strains (Figs. 2B, C). As expected for ATP-mediated Ca2+ release, sample treatment with EGTA, a cell impermeant chelator of extracellular Ca2+ did not decrease the percent of Ca2+ responding cells triggered by wild-type EPEC or the espC mutant (Fig. S2A). The frequency of responses per cell was also not inhibited and even appear to increase upon EGTA-treatment in cells challenged with wild-type EPEC and the espC mutant (Figs. S2B, C). In control experiments, Suramin treatment did not affect actin pedestal structures induced by these strains (Fig. S3).

EPEC-induced Ca2+ responses are elicited by ATP released by the T3SS translocon

HeLa cells were loaded with the fluorescent indicator Cal-520 or with 200 μM suramin for 30 minutes, challenged with the indicated bacteria and subjected to live-cell Ca2+ imaging for a 60 min-duration at a frequency of one acquisition every 10 seconds. A, Representative traces of Ca2+ variations in single cells. The black arrowheads indicate the time of bacterial challenge. The blue arrowheads indicate stimulation with 3 µM histamine. B, Percent of cells exhibiting Ca2+ responses (N = 3, n > 66). C, Average number of responses per cell. Low MOI: 10 bacteria / cell. High MOI: 50 bacteria / cell. Bar: mean. (N = 3, n > 30). Mann-Whitney test. **: p < 0.01; ***: p < 0.001; ****: p < 0.0001.

Together, these results suggest that Ca²⁺ responses induced by EPEC are mediated by ATP released in the extracellular medium via pores formed by the T3SS translocon and are down-regulated by EspC.

EPEC induces coordinated Ca²⁺ responses from single IP3R clusters

We previously showed that Shigella induced local Ca2+ responses dependent on the T3SS and Ca2+ release (Tran Van Nhieu et al., 2013), suggesting that insertion of the Type III translocon was responsible bacterial-induced local Ca2+ signals. We therefore set up to investigate whether EPEC could also trigger T3SS-dependent local Ca2+ responses.

To explore this, we performed rapid Ca2+ imaging at a frequency of 57 ms acquisition per frame to sample elementary Ca2+ release events. As shown in Fig. 3, by performing high speed Ca2+ imaging, we detected fast Ca2+ increases associated with cell challenge with EPEC. These fast Ca2+ responses did not correspond to other responses previously reported since they involved the whole cell or a large cell area but show a small amplitude and fast dynamics usually associated with local Ca2+ responses (Fig. 3B). As observed for the ATP-dependent responses shown in Fig. 2, the espC mutant triggered a higher percent of Ca2+ responding cells than wild-type EPEC (Fig. 3C). Also, the response amplitude was higher for the espC mutant with an average amplitude corresponding to 7.7 ± 0.4 % (mean ± SEM) of the maximal agonist response, compared to 5.4 ± 0.4 % (mean ± SEM) for wild-type EPEC (Fig. 3E). These responses occurred repeatedly at a high frequency of up to 4.5 responses per minute during several minutes following bacterial challenge (Figs. 3D and 3F) and were dependent on Type III Secretion, as evidenced by the lack of response in cells infected with the ΔescN strain (Figs. 3B and 3C). EPEC-induced fast Ca2+ responses were dependent on Ca2+ release since they were inhibited by U73122, a PLC inhibitor (Figs. 3C and 3D). These similarities with the EPEC-induced eATP-dependent Ca2+ responses suggested that the atypical fast responses were triggered by low amounts of ATP released in the extracellular medium following insertion in host cell plasma membranes of discrete numbers of EPEC T3SS translocons. Consistently, these atypical EPEC-induced fast responses were abolished in the presence of the ATP receptor inhibitor Suramin (Figs. 3C, D).

EPEC induces rapid and Coordinated Elementary Ca2+ Responses

HeLa cells were loaded with the fluorescent indicator Cal-520, challenged with the indicated bacteria and subjected to high speed Ca2+ imaging at a frequency of one acquisition every 57 ms for a duration of 110 seconds. A, Representative time-serie of pseudocolored fluorescent micrographs of cells challenged with wild-type EPEC. The numbers indicate the elapsed time in ms from an arbitrarily determined origin. Scale bar = 10 µm. B, D, traces of Ca2+ variations in 2 subcellular regions of the same cell. B, WT: traces corresponding to the regions depicted in the image 0 of panel A. The arrowheads point to the Ca2+ responses shown in Panel A with the corresponding color. C, Percent of cells exhibiting Ca2+ responses (N > 3, n > 256). + Suramin: treatment with 200 μM Suramin. +U73122: treatment with 10 μM U73122. E, Response amplitude expressed as a percent of the maximal response amplitude induced by treatment with 3 μM histamine (N = 3, n > 20). F, average number of responses per cell. C, E, Bar: mean. Mann-Whitney test. **: p < 0.01; ****: p < 0.0001. F, High speed Ca2+ imaging was performed every 5 min for 110 seconds following infection with the indicated bacterial strain as depicted the scheme. The average number of responses per cell is indicated (N > 3, n > 29).

Together, these results show that at the onset of infection, EPEC induces an atypical pattern of fast Ca2+ responses involving the whole or large area of the cell, likely resulting from low ATP levels released by the insertion of a discrete number of translocons in host cell membranes.

EPEC-induced fast Ca2+ responses are triggered by low ATP levels

Previous studies have described local Ca²⁺ increases triggered by sub-maximal agonist concentrations and leading to limited IP3-mediated Ca2+ release. These local Ca²⁺ responses are typically small, of short durations and localized to subcellular regions. Among these, the so-called “Blips” correspond to elementary events of opening of a single IP3 receptor channel usually lasting between 50 and 100 ms, whereas “Puffs” involve the synchronized activation of multiple IP3 receptor channels in localized clusters and last several hundreds of ms (Swillens et al., 1999). In contrast to these described local Ca2+ signals, EPEC-induced fast and small responses could occur throughout the cell, suggesting the coordination of elementary responses over large cell area. Since our findings suggested that these responses were elicited by minute amounts of eATP released by a discrete number of T3SS translocons, we investigated whether low ATP levels could elicit similar responses.

HeLa cells treated with 150 nM ATP showed Ca2+ responses that were indistinguishable from fast responses elicited by EPEC, with an average percent of Ca2+ responding cells of 61.2 ± 5.8 (median ± SEM) and a frequency of 3.9 responses per cell over 60 seconds (Figs. 4A, C and S4A, B). These fast Ca2+ responses had an amplitude that did not exceed 10 % of the maximal agonist response, and occurred over several minutes (Fig. 4A). As observed for EPEC, fast Ca2+ responses induced by low ATP levels involved the whole cell or large cell area encompassing the nuclear and perinuclear area and corresponding to at least 30 % of the cell area as illustrated in Fig. 4B. In this large area, all ROIs corresponding to 1 square micron showed a superimposable profile, as illustrated by traces in Fig. 4C. More detailed scrutinizing showed that in their initial mounting phase, these fast Ca2+ responses were created by the opening of discrete clusters involving an area of ca. 0.04 μm2 that had a transient activity, or possibly were highly mobile, since they were seldom detected at a similar location for three consecutive 22 ms acquisition frames (Fig. S5). These discrete clusters showed similar Ca2+ kinetics suggesting the coordination of Ca2+ release of single IP3R clusters throughout the area that we will hereafter termed CCRICs for “Coordinated Ca2+ Responses from IP3R Clusters”. Treatment with BAPTA-AM to chelate intracellular Ca2+ led to a complete inhibition of CCRICs (N = 3, n > 150 cells; Fig. S4A). Ca2+ responses could still be detected upon cell treatment with EGTA-AM consistent with its lower kon rate for Ca2+, but with a significant inhibition of the percentage of Ca2+ responding cells as well as of the frequency of responses per cell, suggesting that coordination could occur via Ca2+ diffusion and Ca2+-induced Ca2+ release (Fig. S4A-C). Local responses elicited by these ATP concentrations could also be detected often occurring at the cell periphery (Figs. 4B, red region and 4C, red arrow). While the CCRICs area and the peripheral puff area showed spatially segregated Ca2+ responses at the onset of the elicitation (Fig. 4C, peaks 1, 3 and 2, 4), this spatial segregation appeared to blur over the duration of the experiment with peaks integrating responses from the whole cell (Fig. 4C, green and red dashed lanes).

EPEC-induced Coordinated Elementary Ca2+ Responses are reproduced by low ATP levels

A-C, HeLa cells were loaded with the fluorescent indicator Cal-520, challenged with 150 nM ATP and subjected to Ca2+ imaging. Image acquisition every 2 s (A) or 22 ms (B, C). A, Traces of Ca2+ variations corresponding to a single cell (red trace), or subcellular regions within the same cell (inset). A, arrowhead: challenge with 2 μM ATP. B, Time serie of fluorescent micrographs pseudocolored using the “glow” Fiji lookup table, where the blue pixel correspond to an arbitrarily set threshold value. The numbers indicate the elapsed time in seconds. Blue: high intensity pixels showing the large top cell area with CCRICs and the local lower puff area. Scale bar = 10 µm. C, Traces corresponding to Ca2+ variations in the subcellular regions depicted in Panel B. The responses are labelled 1-4, with the response 1 corresponding to the puff (Panel B, red ROI) impulsing the response 3 in the same region. Responses 2 and 4 correspond to CCRICs in Panel B, blue and green ROIs. Note the diffusion of the responses from the initial release area in other area inferred from the dampening of the response amplitude. The dashed red and green lines point at the aggregation of responses throughout the cell.

CCRICs are coordinated by the rapid diffusion of Ca2+ at low concentrations in cell area with a high density of IP3 clusters

In Fig. 5, we used modeling to further investigate the mechanism of coordination of these fast Ca2+ responses. Based on our previous studies (Voorsluijs et al., 2019; Ornelas-Guevara et al., 2023), the model provides a fully stochastic spatial description of Ca2+ release dynamics from IP3R clusters in a two-dimensional representation of a HeLa cell. The simulation domain extends on 10 × 10 µm2 and is discretized into a 20 × 20 grid of compartments (0.5 × 0.5 µm2 each), each representing a cytosolic subvolume of 10-16 L. Each compartment contains at most one cluster of IP3R, whose dynamics is described as a whole. Besides, cytosolic Ca2+ concentration can also vary because of uptake by SERCA, release by a leak or diffusion. A full description of the model can be found in the Supplementary Methods.

Modeling of Coordinated Elementary Ca2+ Responses

Top, Ca2+ variations in subcellular area within a single cell are represented in pseudocolor. Shown are the maximum values of Δ[Ca2+]/[Ca2+]b reached in each compartment during a 60s simulation. Empty white squares: IP3R clusters. Graphs, Traces correspond to Ca2+ variations in the region with the matching color. Colored arrows: Ca2+ response due to the activation of an IP3 cluster in the region with the matching color. Colored arrowhead and dashed red and blue lanes: Ca2+ variations due to the diffusion of a Ca2+ response from or nearby to the region with the matching color. Black arrows: Ca2+ response due Ca2+-activated Ca2+ release. A, low density of IP3R clusters with local responses detected. B, C, Empty black box: area with a high density IP3R clusters. C, similar to B, but following IP3R cluster sensitization due to increased Ca2+ responses.

In Fig. 5A, low ATP levels lead to low IP3 levels activating a limited number of IP3 clusters, opening stochastically and releasing small amounts of Ca2+. In a given area of the cell, the Ca2+ variation integrates Ca2+ release from clusters within this area, as well as Ca2+ diffusing from or to other cell area. For a given response, the initially firing cluster is contained in an area characterized by the highest Ca2+ peak amplitude (Fig. 5A, blue and green arrows) that dampens in a distal area (Fig. 5A, blue and green arrowheads). If the density of IP3R clusters is low, as expected for ER compartment at the cell periphery, the spatial segregation of the initial firing cluster and resulting Ca2+ diffusion to other area is clearly detected (Fig. 5A). In instances, however, the model predicts temporally coordinated responses of similar amplitude, suggesting Ca2+-induced Ca2+ release from secondary clusters (Fig. 5A, black arrow). For both type of Ca2+ dynamics, a large value of the Ca2+ diffusion coefficient of 100 μm2 / s is a key parameter that needs to be taken into account in the model. While it is generally admitted that Ca2+ diffuses very slowly due to the Ca2+ buffers in the cell (∼30 μm2/s), the low levels of Ca2+ released by CCRICs may not be subjected to the diffusion limitations observed at higher Ca2+ levels, because of the relative moderate affinity of buffers for Ca2+.

If the IP3R cluster density is high, as expected in the large perinuclear and nuclear area corresponding to the bulk of the ER, the coordination between individual clusters is very fast (Fig. 5B). As a result, the identification of initial firing clusters goes beyond the technical capacities of the imaging set-up, and ROI within this area show comparable profiles of fast Ca2+ responses (Fig. 5B). Upon prolonged incubation with increasing Ca2+ responses and sensitization of IP3R clusters, the coordination of responses linked to Ca2+-induced Ca2+ release becomes predominant throughout the cell (Fig. 5C).

These findings indicate that the novel Ca²⁺ response pattern is not exclusive to EPEC infection and can be replicated by low levels of eATP, suggesting a broader physiological relevance. From an immunological perspective, CCRICs may therefore play a critical role in various signaling pathways during bacterial infection.

Low eATP levels dampen NF-κB activation

eATP is a well-characterized danger signal contributing to the elicitation of pro-inflammatory signals in various tissues in response to infections (Savio et al., 2018). Previous studies linked intracellular Ca²⁺ signaling and NF-κB activation, a key transcription factor that triggers inflammatory responses (Smedler et al., 2014). We therefore set up to investigate the effects of CCRICs triggered by low eATP levels on NF-κB activation, by performing Western blot analysis against the phosphorylated forms of IκBα (p-IκBα) and P65 (p-P65).

As shown in Figs. 6A and 6B, in control samples, TNF-α induced IκB-α phosphorylation peaking 10 minutes following challenge. In contrast, in the presence of low ATP levels, IκB-α showed a delayed phosphorylation with a 2.1 fold decrease at 10 minutes post-challenge (Figs. 6A and 6B). Consistently, the rates of IκB-α degradation were also slower in the presence of ATP relative to control (Figs. 6A and 6C). As expected from the IκB-α results, TNF-α induced the phosphorylation of the NF-κB P65 subunit and ATP led to a delay and decrease in P65 phosphorylation (Figs. 6D and 6E).

Low ATP levels dampen NF-kappaB activation

HeLa cells were stimulated with 10 ng/ml TNF-α alone or in the presence of 150 nM ATP or 20 μM BAPTA-AM (F-H). At the indicated time points, cell lysates were analyzed by Western blot using the indicated antibodies. A, D, F, G, Representative blots. B-E, H, Densitometry analysis of the indicated antibody signal normalized to that of HSP90 (B, C, F) or total P65 (E). p-P65: anti-phospho P65 antibody. P-IκB: anti-phospho IκB antibody. ANCOVA test. *: p < 0.05; **: p < 0.01; ***: p < 0.001.

In control experiments, we did not detect differences in P65 phosphorylation in response to TNF-α stimulation when cells were treated with BAPTA-AM to chelate intracellular Ca2+ (Figs. 6F and 6H). However, cell treatment with BAPTA-AM prevented the dampening of P65 phosphorylation triggered by low ATP levels (Figs. 6G and 6H), suggesting that CCRICs down-regulated TNF-α -induced NF-κB activation.

Low eATP levels down-regulates NF-κB activation through Ca2+-dependent O-GlcAcylation

We next investigated how CCRICs could regulate NF-κB activation. O-linked β-N-acetylglucosamine (O-GlcNAc) transferase (OGT) was reported to regulate NF-κB signaling by post-translationally modifying the p65 subunit (Ruan et al., 2017). Interestingly, OGT is regulated by Ca2+ signaling, suggesting that CCRICs could affect NF-κB activation via O-GlcNAcylation.

As shown in Fig. S6, TNF-α in the presence of 150 nM eATP stimulated the levels of O-GlcNacylation, specifically for proteins with an apparent molecular weight superior to 100 kDa, that was not observed with TNF-α alone. To further investigate the effects of low eATP levels on NF-κB O-GlcNacylation, we performed immunoprecipitation of P65 RelA on lysates of cells stimulated for 12 minutes with TNF-α alone or co-stimulated with TNF-α and 150 nM eATP. As shown in Fig. 7, TNF-α induced increased O-GlcNacylation of P65 relative to non-stimulated cells, but this increase was inhibited by low eATP levels. Inhibition of P65 O-GlcNacylation by eATP was Ca2+ dependent, since it was not observed in the presence of BAPTA-AM (Fig. 7).

Low ATP levels down-regulate NF-kB O-GlcNAcylation in a Ca2+-dependent manner

HeLa cells were stimulated with 10 ng/ml TNF-α alone or in the presence of 150 nM ATP or 20 μM BAPTA-AM for 12 min. Cell lysates were subjected to P65 immunoprecipitation. A, Representative blots with the indicated antibodies. IP: immunoprecipitates; L: total cell lysates. B, Densitometry analysis of the O-GlucNAc signal in P65 immunoprecipitates normalized to that of TNF-α alone. Mann-Whitney test. N = 4. *: p < 0.05. ns: not significant.

The results indicate that low eATP levels inhibit O-GlcNacylation of P65 induced by TNF-α in a Ca2+-dependent manner and suggest that differential O-GlcNacylation of P65 relative to higher molecular weight proteins.

Discussion

We report here the characterization of CCRICs, corresponding to yet undescribed Ca2+ responses showing the rapid kinetics and small amplitude of local Ca2+ responses but coordinated over large cell area. We found that CCRICs implicate the rapid coordination of IP3R clusters in large cell area, challenging established concepts in the Ca2+ signaling field.

Ca²⁺ diffusion in the cytosol is regulated by mobile and immobile Ca²⁺-binding proteins acting as buffers and forming localized microdomains with steep Ca²⁺ concentration gradients. At the mouth of an open IP₃R channel, Ca²⁺ concentration can reach 100 μM, while just 1–2 μm away, it may drop below 1 μM. Diffusion is restricted by Ca2+ buffers with KD’s that are generally lower than this concentration. Consequently, Ca²⁺ signaling is generally admitted to be spatially restricted, typically influencing regions within approximately 5 μm of the release site (Foskett et al., 2007). The distribution of Ca²⁺-binding proteins and the spatial arrangement of release channels allow IP₃R-mediated [Ca²⁺]i signals to exhibit diverse spatial and temporal properties, making this system highly adaptable (Vandeput et al., 2007). High-resolution optical imaging of fluorescent Ca²⁺ indicators in intact cells indicates that IP₃-mediated [Ca²⁺]i signals are structured levels (Foskett et al., 2007). At low IP₃ levels, individual IP₃R opens stochastically at discrete release sites, causing localized elevations in cytoplasmic [Ca²⁺]. At higher IP₃ levels, Ca²⁺ release spreads between IP₃R clusters, propagating waves that travel at tens of microns per second, coordinating intracellular signaling ultimately leading to a global Ca²⁺ response. In reference models describing intracellular Ca2+ dynamics, cell regions with a high IP3R density initiate the Ca2+ response from which Ca2+ waves propagate with a diffusion coefficient of 10 – 30 μm2/s (Falcke et al., 2003).

In contrast to these described global and local Ca²⁺ responses, we found CCRICs to be highly temporally coordinated over large area, suggesting the fast diffusion of Ca2+ and propagation of Ca2+ by Ca2+-induced Ca2+ release. While challenging generally admitted concepts on the poor diffusion of Ca2+, this view is fully supported by our theoretical modeling implicating the fast diffusion of Ca2+, with a diffusion coefficient of at least 100 μm2/s that can be expected at low cytoplasmic [Ca²⁺]. Indeed, at these low Ca²⁺ concentrations not exceeding a few hundreds nM, the majority of Ca²⁺ buffers are not expected to efficiently bind to Ca²⁺ and to significantly interfere with Ca²⁺ diffusion because of their relative low affinity.

We found that CCRICs implicate a large cell area including the nuclear and perinuclar area, while local Ca²⁺ responses were detected at small area of the cell periphery. The large CCRIC area likely involves the bulk of the endoplasmic reticulum, while peripheral area may contain smaller ER compartments. In our model considering fast Ca2+ diffusion when buffers are far from saturation, the higher density of IP3R clusters in the CCRIC area accounts for the high coordination of the responses, relative to the lack of coordination in peripheral area where lower IP3R density is expected.

We showed that a low dose of eATP triggering CCRICs delayed and dampened NF-κB activation linked to a reduction in O-GlcNAcylation of the NF-κB p65 subunit. One major open question is the mechanism by which CCRICS down-regulate NF-κB activation. NF-κB activity can be modulated by O-GlcNAcylation, a reversible glycosylation modification catalyzed by O-GlcNAc transferase (OGT) and O-GlcNAcase (OGA) (Liu & Ramakrishnan, 2021). Previous studies demonstrated that Ca²⁺ signals activate Ca²⁺-regulated enzymes like CaMKII, which in turn phosphorylates and activate OGT, promotes O-GlcNAcylation (Ruan et al., 2017). In other studies, OGT-mediated O-GlcNAcylation could modulate NF-κB signaling pathway (X. Dong et al., 2023). O-GlcNAcylation of P65 could also inhibit its interaction with IκB-α, promote p65 nuclear translocation and increase NF-κB transcriptional activity (Liu & Ramakrishnan, 2021). Reduced O-GlcNAcylation of P65 at residues S550 and S551 was shown to result in decreased NF-κB activation and nuclear translocation (Motolani et al., 2023). These findings are in line with our results suggesting that CCRICs elicited by low-level eATP, downregulate p65 O-GlcNAcylation and NF-κB activation, possibly by modulating OGT activity or OGT-p65 interactions. Reduced O-GlcNAcylation of p65 may affect its phosphorylation patterns indirectly, perhaps by altering the interaction of NF-κB with kinases or phosphatases involved in its activation (Özcan et al., 2010).

Our findings indicate that eATP differentially regulate inflammatory signaling pathways in epithelial cells by dampening NF-κB activation at low levels and stimulating its activation at high concentrations. These results extend the key role of eATP from a danger-associated molecular pattern (DAMP) to a fine-tuner of inflammatory responses depending on its concentration.

Materials and Methods

Cell and Bacterial culture

HeLa cells were maintained in Dulbecco’s Modified Eagle Medium (DMEM; Gibco, Thermo Fisher Scientific) supplemented with 10% fetal bovine serum (FBS; Gibco, Thermo Fisher Scientific) at 37 °C in a humidified incubator with 10% CO₂. Wild-type Enteropathogenic Escherichia coli (EPEC WT), ΔescN, and ΔespC strains were cultured in Luria-Bertani (LB) broth at 37 °C with kanamycin at a final concentration of 15 μg/ml in a shaking incubator. All strains were transformed with the pmCherry-N1 plasmid, which encodes red fluorescent protein and carries an ampicillin resistance gene for selection. Where applicable, ampicillin (100 μg/mL) and kanamycin were included to maintain resistance markers. The red fluorescence enabled visualization of the bacteria during downstream analyses.

EPEC infection of HeLa cells

HeLa cells were seeded in 6-well plates at a density of 4.5 × 10⁵ cells per well one day before infection. EPEC WT, ΔescN, and ΔespC strains grown in the exponential phase were resuspended and primed for 5 hours before challenging HeLa cells in DMEM medium. Cells were challenged with bacteria at an OD600 = 0.2 (Low MOI) or 0.8 (High MOI).

Ca2+ imaging

HeLa cells were seeded onto 25 mm-diameter glass coverslips. Cells were preloaded with the fluorescent indicator dye Cal-520 (AAT #21130) for 30 minutes at room temperature, followed by two PBS washes and one time with DMEM. And placed coverslips, imaging was performed in an observation chamber in DMEM without phenol red, supplemented with 25 mM HEPES. Following a 3-minute baseline acquisition, add the required bacteria strains and OD600 into the chamber. Following a 10-minute incubation at room temperature to allow bacterial attachment. Imaging was then carried out at 35°C using a Nikon Eclipse TE200 inverted fluorescence microscope with a 60× objective for 1h of infection. Fluorescence signals were acquired 485 nm with excitation and 535 nm emission parameters. Image control and data acquisition were managed by Simple32 software (Compix Inc.), depending on the experiment requirement, imaging at 22ms, 57ms or 10 sec acquisition intervals. 3 μM of Histamine and 2 μM of ionomycin were applied to check the ability of cells to show calcium response. Images were captured using a CMOS camera (Hamamatsu) and analyzed using the same software.

Immunofluorescence analyses

Cells were washed three times with PBS and fixed with 3.7% paraformaldehyde (PFA), permeabilized with 0.1% Triton X-100 for 5 minutes and washed with PBS. Blocking was performed using 3% FBS in PBS for 30 minutes at room temperature. Cells were incubated with primary antibodies for 2 hours: mouse anti-phalloidin (1:200; Fischer Scientific, #17511176). Secondary antibodies (anti-mouse), along with DAPI were applied for 1 hour. Samples were mounted using Dako mounting medium (Agilent) and imaged using a Nikon Ti2 confocal microscope equipped with a 60× objective and Nikon acquisition software.

Modelling

We develop a fully stochastic spatial model to simulate Ca2+ release dynamics from IP3R clusters in a two-dimensional representation of a HeLa cell. The simulation domain measures 10 × 10 µm2 and is discretized into a 20 × 20 grid of compartments (0.5 × 0.5 µm2 each), each representing a cytosolic subvolume of 10-16 L. Ca2+ exchange between the Endoplasmic Reticulum (ER) and the cytosol occurs through IP3R-mediated release, SERCA uptake, and ER Ca2+ leak, following the framework by Voorsluijs et al., 2019 and Ornelas-Guevara et al., 2023. Ca2+ diffusion is implemented stochastically as a kinetic process between adjacent compartments (Kraus et al., 1996), with a diffusion coefficient of 100 µm2/s to reflect moderate endogenous buffering. Each IP3R cluster functions as a single unit with four possible states: Open (O), Closed (C) and two Inhibited states (i1, and i2), transitioning in response to local [Ca2+] and [IP3]. This phenomenological description captures key characteristics of Ca2+ puffs and their transition to global signals via Ca2+ diffusion and Calcium Induced Calcium Release.

We perform all simulations using the Gillespie algorithm, where each event, reaction or diffusion, is selected stochastically based on its propensity. See Supplementary Methods for complete model equations and parameter values.

Western Blot analysis

To obtain total cell extracts, cells were lysed in sample buffer 1x (62.5 mM Tris pH=8, 2% SDS, 10% glycerol, 0.05% bromophenol blue, 5% β-mercaptoethanol) and boiled at 95°C for 5 minutes. Proteins from total lysates were separated by SDS PAGE and transferred to nitrocellulose membrane (0.45μM AmershamTM ProtranTM). Western Blot analysis was performed according to standard procedure using the following primary antibodies: IkappaBalpha (OZYME, 9424S), Phospho-IκB-α (OZYME, 9424S), NF-kappaB p65 (OZYME, 9246S), Phospho-NF-kappaB p65 (OZYME, 3033S), HSP90 (Santa Cruz Biotechnologies sc-13119). HRP-conjugated anti-mouse (Cytiva) and anti-rabbit (Sigma) were used as secondary antibodies.

Immunoprecipitation assays

HeLa cells were seeded in 150 cm2 dishes (7.4 × 10⁶ cells/dish) and cultured in DMEM supplemented with 10% FBS. Cells were pretreated with 20 µM BAPTA for 30 minutes at room temperature where indicated, then stimulated with 10 ng/mL TNF-α alone or in combination with 150 nM ATP for an additional 12 minutes. After treatment, cells were washed with ice-cold PBS containing 1 mM NaF and 1 mM Na₃VO₄, lysed in ice-cold lysis buffer (50 mM Tris-HCl pH 7.5, 0.5% Triton X-100, 100 mM NaCl, 1 mM DTT, protease inhibitor cocktail without EDTA), and incubated on a rotating wheel at 4°C for 1 hour. Lysates were clarified by centrifugation at 13,000 rpm for 30 minutes at 4°C. Supernatants were incubated with 5 µL of anti-NF-kB p65 antibody (Abcam, ab16502) for 2 hours at 4°C with rotation, followed by overnight incubation with pre-equilibrated protein A/G beads. Immunocomplexes were washed once with lysis buffer and twice with PBS, then eluted in Laemmli buffer by boiling at 100°C for 5 minutes. Input and IP samples were analyzed by Western blotting using antibodies against O-GlcNAc (Abcam, ab2739), anti-NF-kB p65 (OZYME, 9246S), and Actin (OZYME, 4967).

Statistical Analysis

All quantitative data are presented as mean ± SEM from at least three independent experiments. Statistical significance was assessed using unpaired two-tailed Student’s t-tests with unequal variance, unless otherwise specified. GraphPad Prism 7 (GraphPad Software) was used for statistical analysis, and p-values < 0.05 were considered statistically significant.

eLife Assessment

This study reports important advances in our understanding of how enteropathogenic E. coli (EPEC) interacts at the intestinal interface. Solid data describe a novel model of spatially coordinated calcium signaling to modulate NF-kB activation; additional data and clarification of methods would improve the strength of these conclusions. These findings, which integrate imaging, genetics, and computational modeling, provide a new way to consider host-pathogen interactions in EPEC infections that may lead to improved therapies.

Reviewer #1 (Public review):

Summary:

In their article, Guo and coworkers investigate the Ca²⁺ signaling responses induced by Enteropathogenic Escherichia coli (EPEC) in epithelial cells and how these responses regulate NF-κB activation. The authors show that EPEC induces rapid, spatially coordinated Ca²⁺ transients mediated by extracellular ATP released through the type III secretion system (T3SS). Using high-speed Ca²⁺ imaging and stochastic modeling, they propose that low ATP levels trigger "Coordinated Ca²⁺ Responses from IP₃R Clusters" (CCRICs) via fast Ca²⁺ diffusion and Ca²⁺-induced Ca²⁺ release. These responses may dampen TNF-α-induced NF-κB activation through Ca²⁺-dependent modulation of O-GlcNAcylation of p65. The interdisciplinary work suggests a new perspective on calcium-mediated immune response by combining quantitative imaging, bacterial genetics, and computational modeling.

Strengths:

The study provides a new concept for host responses to bacterial infections and introduces the concept of Coordinated Ca²⁺ Responses from IP₃R Clusters (CCRICs) as synchronized, whole-cell-scale Ca²⁺ transients with the fast kinetics typical of local events. This is elegantly done by an interdisciplinary approach using quantitative measurements and mechanistic modelling.

Weaknesses:

(1) The effect of coordination by fast diffusion for small eATP concentrations is explained by the resulting low Ca2+ concentration that is not as strongly affected by calcium buffers compared to higher concentrations. While I agree with this statement on the relative level, CICR is based on the resulting absolute concentration at neighboring IP3Rs (to activate them). Thus, I do not fully agree with the explanation, or at least would expect to use the modelling approach to demonstrate this effect. Simulations for different activation and buffer concentrations could strengthen this point and exclude potential inhibition of channels at higher stimulation levels.

In this respect, I would also include the details of the modelling, such as implementation environment, parameters, and benchmarking. The description in the Supplementary Methods is very similar to the description in the main text. In terms of reproducibility, it would be important to at least provide simulation parameters, and providing the code would align with the emerging standards for reproducible science.

(2) Quantitative characterization of CCRICs:

The paper would benefit from a clearer definition of the term CCRICs and quantitative descriptors like duration, amplitude distribution, frequency, and spatial extent (also in relation to the comment on the EGTA measurements below). Furthermore, it remains unclear to me whether CCRICs represent a population of rapidly propagating micro-waves or truly simultaneous events. Maybe kymographs or wave-front propagation analyses (at least from simulations if experimental resolution is too bad) would strengthen this point.

(3) Specificity of pharmacological tools:

Suramin and U73122 are known to have off-target effects. Control experiments using alternative P2 receptor antagonists like PPADS or inactive U73343 analogs would strengthen the causal link.

Reviewer #2 (Public review):

Summary:

The authors of this study are trying to resolve how cellular infection by enteropathogenic E. coli (EPEC) subverts cellular signaling pathways to promote infection and dampen immune responses. Specifically, alteration in calcium dynamics has been evidenced in the prior literature as a potential initiator of these adaptations, and this study provides ideas and mechanistic detail as to how cellular calcium dynamics may be subverted by pathogens.

Strengths:

The clear strengths of this paper relate to the new ideas inherent in the proposed hypothesis and their support from the experimental approaches used. Overall, the proposed work provides new ideas in this area, which will benefit from further investigation. Certainly, this is an interesting and challenging paradigm to pick apart mechanistically, and is important for improving treatments from intestinal infections.

Weaknesses:

Additional insight is needed in three specific areas to convincingly support the conclusions drawn by the authors. These three areas are: first, a better description of the infection-associated calcium signals. Second, a mechanistic definition of the relevant purinoceptors versus other pathways to increase cellular calcium. Third, an effort to show that the proposed pathways have relevance in a polarized epithelial cell.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public review):

Summary:

In their article, Guo and coworkers investigate the Ca²⁺ signaling responses induced by Enteropathogenic Escherichia coli (EPEC) in epithelial cells and how these responses regulate NF-κB activation. The authors show that EPEC induces rapid, spatially coordinated Ca²⁺ transients mediated by extracellular ATP released through the type III secretion system (T3SS). Using high-speed Ca²⁺ imaging and stochastic modeling, they propose that low ATP levels trigger "Coordinated Ca²⁺ Responses from IP₃R Clusters" (CCRICs) via fast Ca²⁺ diffusion and Ca²⁺-induced Ca²⁺ release. These responses may dampen TNF-α-induced NF-κB activation through Ca²⁺-dependent modulation of O-GlcNAcylation of p65. The interdisciplinary work suggests a new perspective on calcium-mediated immune response by combining quantitative imaging, bacterial genetics, and computational modeling.

Strengths:

The study provides a new concept for host responses to bacterial infections and introduces the concept of Coordinated Ca²⁺ Responses from IP₃R Clusters (CCRICs) as synchronized, whole-cell-scale Ca²⁺ transients with the fast kinetics typical of local events. This is elegantly done by an interdisciplinary approach using quantitative measurements and mechanistic modelling.

Weaknesses:

(1) The effect of coordination by fast diffusion for small eATP concentrations is explained by the resulting low Ca2+ concentration that is not as strongly affected by calcium buffers compared to higher concentrations. While I agree with this statement on the relative level, CICR is based on the resulting absolute concentration at neighboring IP3Rs (to activate them). Thus, I do not fully agree with the explanation, or at least would expect to use the modelling approach to demonstrate this effect. Simulations for different activation and buffer concentrations could strengthen this point and exclude potential inhibition of channels at higher stimulation levels.

We fully agree that CICR is determined by the local Ca2+ concentration at each IP3R cluster, not by a global cytosolic average. In our stochastic model, IP3 R clusters are represented as phenomenological entities at discrete spatial sites. Each cluster senses the local Ca2+ concentration at its position, and its stochastic gating depends only on this local [Ca2+] and on [IP3]. Buffers are not included explicitly. Instead, we use an effective Ca2+ diffusion coefficient Deff, which accounts for the effect of endogenous Ca2+ buffers. To reproduce the coordinated low-amplitude Ca2+ responses observed experimentally, we found that we had to use Deff = 100 µm2/s. In the supplementary analysis, we show that an effective diffusion coefficient of this order is indeed plausible for a realistic mixture of mobile and immobile Ca2+ buffers (Supplementary Note 2. Figure 1).

In the revised manuscript, we now provide a supplementary analysis (Supplementary Note 2) to justify this choice. Using an equation to compute the effective diffusion coefficient considering a plausible mixture of mobile and immobile buffers and an explicit reaction–diffusion model, we show that:

- The effective diffusion coefficient of Ca2+ becomes Ca2+ dependent, and

- There exists a regime in which low-amplitude Ca2+ elevations are characterized by an effective diffusion coefficient of Deff = 100 µm2/s and a larger spatial extent than higher-amplitude transients (Supplementary Note 2. Figure 1).

Thus, the value of Deff used in the cluster model is quantitatively consistent with classical buffering theory and with plausible cytosolic buffer mixtures. This provides a mechanistic basis for the observation that small-amplitude, short-lived events can nevertheless produce coordinated signals with large spatial extent and, occasionally, almost immediate activation of IP3R clusters at distant locations in both simulations and experiments.

In this respect, I would also include the details of the modelling, such as implementation environment, parameters, and benchmarking. The description in the Supplementary Methods is very similar to the description in the main text. In terms of reproducibility, it would be important to at least provide simulation parameters, and providing the code would align with the emerging standards for reproducible science.

We apologize for the lack of details of the modelling in the previous submission. In this revised version, we are providing with a full description of the model in the Supplementary Information, Note1.

To address the reviewer’s request for simulations at different activation levels, we now show an additional simulation in which [IP3] is higher (0.1 µM, constant in time and space) and Deff is set to 40 µm2/s (Supplementary Note 3). This lower effective diffusion coefficient is consistent with the stronger buffering and reduced Ca2+ mobility expected for higher-amplitude signals. In this case, the same phenomenological cluster model generates a global Ca2+ response with larger amplitude and longer duration, rather than a loss of activity due to excessive inhibition ((Supplementary Note 3, Figure 1, left panel). The Supplementary Note 3. Figure 1, right panel shows the 2D cell geometry, where dots indicate the random positions of IP3R clusters whose behavior is described by our phenomenological cluster model.

(2) Quantitative characterization of CCRICs:

The paper would benefit from a clearer definition of the term CCRICs and quantitative descriptors like duration, amplitude distribution, frequency, and spatial extent (also in relation to the comment on the EGTA measurements below). Furthermore, it remains unclear to me whether CCRICs represent a population of rapidly propagating micro-waves or truly simultaneous events. Maybe kymographs or wave-front propagation analyses (at least from simulations if experimental resolution is too bad) would strengthen this point.

We agree and completed the description of the CCRICs by adding:

In the Results section, p. 8, l. 27:

“…with a duration of 2.1 ± 1.0 sec (mean ± SEM) (N = 4, 128 responses)”. p. 9, l. 13:

“In rare instances (less than 3%), typical local “Puff” responses elicited by these ATP concentrations could also be detected often occurring at the cell periphery (Figs. 4B, red region and 4C, red arrow; Fig. S6D, blue trace) (N > 20, cells > 500). As expected from the small concentrations of Ca2+ released at puff sites, no increase in cytosolic Ca2+ was detected in a distal cell region (Fig. S6D, top), indicating that isotropic Ca2+ diffusion from a puff release site cannot account for Ca2+ increase over large cell area. Puffs could also be detected concomitantly with CCRICs in different ROIs of the same cell (Fig. S6D, bottom). In contrast to puffs, CCRICs often showed responses of comparable amplitude in distal regions over the whole cell (Figs. 4C and S6A, B), suggesting the contribution from IP3R cluster activation by Ca2+-Induced Ca2+ Release (CICR). Within a given cell, the vast majority of CCRICs appeared quasi-synchronized at the fatest acquisition rate of 22 ms / frame that we could achieve. However, in few instances a delay could be detected in the elicitation of a peak in distant region of a cell (Fig. S6C). These observations suggest that the quasi-synchronization of CCRICs result from the fast diffusion of Ca2+ leading to the activation of IP3R clusters over large cell area, which may be delayed in a some instances. Scrutinizing of CCRICs showed that while their profiles were comparable, the amplitude of these responses varied in different regions of the cell, with often a single 1 µm2 region, likely corresponding the initial firing cluster, showing a prominent amplitude and other regions with smaller amplitude for a given response (Figs. 4B and 4C). For example, in Fig. 4C, the highest amplitude is observed in the red region for peaks 1 and 3, whereas it is observed and in the purple region for peak 2. Thus, for a given CCRIC, the respective contribution of local IP3R cluster activation and isotropic diffusion of Ca2+from other release sites in Ca2+ increase may vary in different regions of the cell”.

In the Discussion section, 2nd sentence p. 12:

“CCRICs showed rapid kinetics with an average duration of ca 2.1 seconds and amplitude corresponding to an increase in Ca2+ cytosolic concentration of a few hundreds nM, seemingly smaller than that of puffs (Fig. S6D), often occurring repeatedly with a frequency of up to 12 CCRICs / min over the whole cell.”

We have tried to clarify the notion of coordination versus synchronization of CCRICs by showing the delay observed in some instances in the elicitation of CCRICs at distal regions of the cell, now illustrated shown in Fig S6C.

(3) Specificity of pharmacological tools:

Suramin and U73122 are known to have off-target effects. Control experiments using alternative P2 receptor antagonists like PPADS or inactive U73343 analogs would strengthen the causal link.

As suggested by the referee, we have performed complementary experiments showing the inhibitory effects of PPADS and absence of effects of U73343 on EPEC-induced Ca2+ responses including CCRICs now shown in the amended Fig. S2.

Reviewer #2 (Public review):

Summary:

The authors of this study are trying to resolve how cellular infection by enteropathogenic E. coli (EPEC) subverts cellular signaling pathways to promote infection and dampen immune responses. Specifically, alteration in calcium dynamics has been evidenced in the prior literature as a potential initiator of these adaptations, and this study provides ideas and mechanistic detail as to how cellular calcium dynamics may be subverted by pathogens.

Strengths:

The clear strengths of this paper relate to the new ideas inherent in the proposed hypothesis and their support from the experimental approaches used. Overall, the proposed work provides new ideas in this area, which will benefit from further investigation. Certainly, this is an interesting and challenging paradigm to pick apart mechanistically, and is important for improving treatments from intestinal infections.

Weaknesses:

Additional insight is needed in three specific areas to convincingly support the conclusions drawn by the authors. These three areas are: first, a better description of the infection-associated calcium signals. Second, a mechanistic definition of the relevant purinoceptors versus other pathways to increase cellular calcium. Third, an effort to show that the proposed pathways have relevance in a polarized epithelial cell.

(1) first, a better description of the infection-associated calcium signals.

We agree and have added a more detailed description of the CCRICs in the results and discussion section, as detailed in response to referee 1, Weakness 2 by adding:

In the Results section, p. 8, l. 27:

“…with a duration of 2.1 ± 1.0 sec (mean ± SEM) (N = 4, 128 responses)”. p. 9, l. 13:

“In rare instances (less than 3%), typical local “Puff” responses elicited by these ATP concentrations could also be detected often occurring at the cell periphery (Figs. 4B, red region and 4C, red arrow; Fig. S6D, blue trace) (N > 20, cells > 500). As expected from the small concentrations of Ca2+ released at puff sites, no increase in cytosolic Ca2+ was detected in a distal cell region (Fig. S6D, top), indicating that isotropic Ca2+ diffusion from a puff release site cannot account for Ca2+ increase over large cell area. Puffs could also be detected concomitantly with CCRICs in different ROIs of the same cell (Fig. S6D, bottom). In contrast to puffs, CCRICs often showed responses of comparable amplitude in distal regions over the whole cell (Figs. 4C and S6A, B), suggesting the contribution from IP3R cluster activation by Ca2+-Induced Ca2+ Release (CICR). Within a given cell, the vast majority of CCRICs appeared quasi-synchronized at the fatest acquisition rate of 22 ms / frame that we could achieve. However, in few instances a delay could be detected in the elicitation of a peak in distant region of a cell (Fig. S6C). These observations suggest that the quasi-synchronization of CCRICs result from the fast diffusion of Ca2+ leading to the activation of IP3R clusters over large cell area, which may be delayed in a some instances. Scrutinizing of CCRICs showed that while their profiles were comparable, the amplitude of these responses varied in different regions of the cell, with often a single 1 µm2 region, likely corresponding the initial firing cluster, showing a prominent amplitude and other regions with smaller amplitude for a given response (Figs. 4B and 4C). For example, in Fig. 4C, the highest amplitude is observed in the red region for peaks 1 and 3, whereas it is observed and in the purple region for peak 2. Thus, for a given CCRIC, the respective contribution of local IP3R cluster activation and isotropic diffusion of Ca2+ from other release sites in Ca2+ increase may vary in different regions of the cell” In the Discussion section, 2nd sentence p. 12:

“CCRICs showed rapid kinetics with an average duration of ca 2.1 seconds and amplitude corresponding to an increase in Ca2+ cytosolic concentration of a few hundreds nM, seemingly smaller than that of puffs (Fig. S6D), often occurring repeatedly with a frequency of up to 12 CCRICs / min over the whole cell.”

We have tried to clarify the notion of coordination versus synchronization of CCRICs by showing the delay observed in some instances in the elicitation of CCRICs at distal regions of the cell, now illustrated shown in Fig S6C.

CRICCs are observed over the whole cell or very large cell area. We agree that this point as well as comparison with previously described puffs needed clarification. We have added the following sentences in the discussion and inserted the seminal Thomas et al. 1999 citation in the references, p. 13, l. 18:

“Consistently, while CRICCs were detected in the vast majority of cells at these very low agonist concentrations, in rare instances, local “puff-like” responses were also detected at the cell periphery. These observations are in contrast to previously described Ca2+ puffs preceding global responses reported to occur preferentially in perinuclear area (Thomas et aL., 1999). These earlier studies, however, involved higher agonist concentrations (1-5 µM ATP) expected to lead to the release of higher IP3 concentrations, which may preferentially stimulate larger IP3R clusters at the perinuclear region because of the higher density of IP3 Rs. In addition, larger IP3 clusters may release higher amounts of Ca2+ for which, as opposed to CCRICs, diffusion would be restrained by Ca2+ buffers thereby favoring the spatial confinement of the response. “

(2) Second, a mechanistic definition of the relevant purinoceptors versus other pathways to increase cellular calcium

We do not believe that CCRICs are specific to EPEC, since they are also elicited by low agonist concentrations. The discrete action of Type III translocons leading to the release of small amounts of extracellular ATP at the onset of EPEC prompted us to perform fast Ca2+ imaging at low agonists concentrations (150 nM ATP, 100 nM histamine now shown in Fig. S4), which to our knowledge, differ from higher agonist concentrations used in all previous studies describing puffs. Our modelling studies support the notion that CCRICs correspond to generic Ca2+ release-dependent responses triggered by low levels of IP3.

We now show inhibition of CCRICs by PPADS, another purinergic receptor antagonist, and extracellular ATP depletion by addition of hexokinase in the extracellular medium in Figs. S4 and S7.

Knocking down ATP receptors represents a challenging task since HeLa cells were shown to express transcripts for most of the described 8 P2Xs and 7 P2Ys purinergic receptors (10.1016/j.bbamem.2009.03.006). Mostly, we do not believe that CCRICs are triggered by a specific ATP receptor and do not expect to see inhibition of CCRICs in single knock-down experiments. Our experimental and modelling studies suggest that CCRICs are not specific to EPEC nor to a particular ATP receptor, but instead correspond instead to generic Ca2+ elicited at low agonist concentrations such as ATP or histamine.

Zhong et al., 2020 indeed previously showed a role for Ca2+ influx mediated by the TRPV2 receptor in EPEC-mediated cell death. However, this influx occurred following 8 hours of cell infection with EPEC. We do not detect significant cell death or Ca2+ influx at the onset of infection corresponding to the 12 hours infection kinetics that we used. Our experiments indicate that CCRICs do not involve Ca2+ influx.

(3) Third, an effort to show that the proposed pathways have relevance in a polarized epithelial cell.

We agree and have performed complementary experiments showing induction of CCRICs by EPEC and eATP in polarized intestinal epithelial cells, now shown in Figure S8.

Recommendations for the authors:

Reviewer #1 (Recommendations for the authors):

(1) Statistical treatment and data presentation:

Some figure legends lack clarity on replicates (n = cells vs N = independent experiments). Timecourse quantifications of p-IκB and p-p65 should include normalized fold-change plots with clear statistical tests.

To clarify, we replaced “n” by “cells”. The number of determinations and independent experiments (N) has been added in the legends to all relevant Figures and Supplementary Figures.

As requested, we now show the p-IκB and p-p65 plots as plots normalized to basal p-IκB and p-p65 levels. We mentioned in legend to Fig. 6 that we used an ANCOVA test showing significance of the effects of eATP on TNF-∝-induced IκB- and p65 phosphorylation.

(2) Clarification on the temperature used in imaging (why measured at 35{degree sign} C)?

We have added the following clarification in the Materials and Methods section p. 14, l. 21:

“Imaging was then carried out at 35°C to allow for bacterial type III secretion, …”

(3) Figure 4A:

The image shows a lower image acquisition interval than every 2s that is stated in the caption.

We apologize for the mistake. The legend to Fig. 4A now reads:

“Image acquisition every 52 ms (A)…”

(4) Figure 4B:

The color of ROIs could be more intense for better identification.

We have replaced the colors of blue and green ROIs, by light cyan and purple ROIs

(5) Figure 4c:

I don't understand the meaning of the dashed lines described by "The dashed red and green lines point at the aggregation of responses throughout the cell" in the caption or in the text.

We apologize for the lack of clarity and have re-written the corresponding text p. 9, l.25 as follows:

“Scrutinization of CCRICs showed that while their profiles were comparable, the amplitude of these responses varied in different regions of the cell, with often a ca 3 µm2 single region, likely corresponding to a source point release, showing a prominent amplitude and other regions with smaller amplitude for a given response (Figs. 4B and 4C). For example, in Fig. 4C, the highest amplitude is observed in the red region for peaks 1 and 3, whereas it is observed and in the purple region for peak 2. Thus, for a given CCRIC, the respective contribution of local IP3R cluster activation and isotropic diffusion of Ca2+ from other release sites in Ca2+ increase may vary in different regions of the cell.”

(6) Figure S4A:

The responses for EGTA are not really pointed out. Are the traces meant to show events?

We have added arrowheads in traces corresponding to ATP + EGTA-AM treatment pointing at “flattened Ca2+ responses”. The Legend to Fig. S4A now includes the sentence: “ATP + EGTA-AM treatment led to an inhibition of Ca2+ responses, associated with small variations in the Ca2+ baseline, that were arbitrarily scored as flattened Ca2+ pseudo-responses (ATP+EGTA-AM, red arrows).”

(7) Figure S5:

Could not identify the purple arrow for the less mobile cluster.

We agree that the former Figure lacked clarity and have remade Figure S5, now Figure S6, with higher magnification of panels with fast acquisition. The previously purple arrows pointing at larger and less mobile clusters are now shown in black in these enlarged panels. The legend has been changed accordingly.

(8) There are some typos and suboptimal formulations throughout the manuscript, such as:

P8: "minute amount" could be changed to low, minor or similar.

“minute” amounts of eATP was replaced by “low amounts of eATP”.

P8: put a "%" to the numbers 61.2 {plus minus} 5.8.

“%” was added.

P16: "manuscript".

Thank you.

Reviewer #2 (Recommendations for the authors):

Suggestions relate to the following three topics.

First, a better description of the infection-associated calcium signals. The authors emphasize throughout the paper that their imaging data challenge established concepts in the calcium signaling field (discussion). I do not see the calcium imaging data explained either with data or textually with sufficient clarity to evaluate this assertion. A start would be a clear description of the characteristics of the EPEC-evoked calcium signals relative to other local and global domains of calcium signaling previously described in HeLa cells. Prior work has shown that PI-coupled agonists evoke local calcium signals that are perinuclear in HeLa cells (PMID: 10660296), but the relationship of EPEC-evoked transients to these previously defined responses is not clear.

We agree and have added a more detailed description of the CCRICs in the results and discussion section, as detailed in response to referee 1, Weakness 2.

Most importantly, it is ambiguous where in the HeLa cell recordings are made. Are these recordings close to the plasma membrane and/or deeper within the cell? The only spatial information is provided in Figure 3A, and these responses are not well described in the text or presented in a way that comparisons can be made to responses from a PI-coupled agonist.

CRICCs are observed over the whole cell or very large cell area. We agree that this point as well as comparison with previously described puffs needed clarification. We have added the following sentences in the discussion and inserted the seminal Thomas et al. 1999 citation in the references, p. 13, l. 18:

“Consistently, while CRICCs were detected in the vast majority of cells at these very low agonist concentrations, in rare instances, local “puff-like” responses were also detected at the cell periphery. These observations are in contrast to previously described Ca2+ puffs preceding global responses reported to occur preferentially in perinuclear area (Thomas et aL., 1999). These earlier studies, however, involved higher agonist concentrations (1-5 µM ATP) expected to lead to the release of higher IP3 concentrations, which may preferentially stimulate larger IP3R clusters at the perinuclear region because of the higher density of IP3Rs. In addition, larger IP3 clusters may release higher amounts of Ca2+ for which, as opposed to CCRICs, diffusion would be restrained by Ca2+ buffers thereby favoring the spatial confinement of the response. “

If I understand the described responses correctly, could not these rapid local responses result from a change in cellular calcium buffering capacity consequent to infection? Are the authors proposing that these responses occur in other cells also, or represent a pathogen-specific signaling mode?

We do not believe that CCRICs are specific to EPEC, since they are also elicited by low agonist concentrations. The discrete action of Type III translocons leading to the release of small amounts of extracellular ATP at the onset of EPEC prompted us to perform fast Ca2+ imaging at low agonists concentrations (150 nM ATP, 100 nM histamine now shown in Fig. S4), which to our knowledge, differ from higher agonist concentrations used in all previous studies describing puffs. Our modelling studies support the notion that CCRICs correspond to generic Ca2+ release-dependent responses triggered by low levels of IP3.

Second, evidence supporting a mechanistic role of ATP comes from prior literature, together with the authors' presented data showing the effects of PLC (to inhibit IP3), pharmacological inhibition (suramin, a non-selective purinoceptor blocker), and the effects of T3SS-deficient mutants (to prevent ATP release). However, there are missing steps here to mechanistically identify how ATP is working. First, does degradation of extracellular ATP (e.g., apyrase) block these responses? Second, given HeLa cells are easily amenable to knockdown approaches, does knockdown of particular ATP receptors, or TRPV2 as suggested in the prior literature, impact the calcium signal dynamics?

We now show inhibition of CCRICs by PPADS, another purinergic receptor antagonist, and extracellular ATP depletion by addition of hexokinase in the extracellular medium in Figs. S4 and S7.

Knocking down ATP receptors represents a challenging task since HeLa cells were shown to express transcripts for most of the described 8 P2Xs and 7 P2Ys purinergic receptors (10.1016/j.bbamem.2009.03.006). Mostly, we do not believe that CCRICs are triggered by a specific ATP receptor and do not expect to see inhibition of CCRICs in single knock-down experiments. Our experimental and modelling studies suggest that CCRICs are not specific to EPEC nor to a particular ATP receptor, but instead correspond instead to generic Ca2+ elicited at low agonist concentrations such as ATP or histamine.

Zhong et al., 2020 indeed previously showed a role for Ca2+ influx mediated by the TRPV2 receptor in EPEC-mediated cell death. However, this influx occurred following 8 hours of cell infection with EPEC.

We do not detect significant cell death or Ca2+ influx at the onset of infection corresponding to the 12 hours infection kinetics that we used. Our experiments indicate that CCRICs do not involve Ca2+ influx.

Third, while the use of HeLa cells provides advantages for imaging and mechanistic assays, the effort to replicate findings in an intestinal cell line would heighten relevance, given the likely importance of cell type and cell polarity on the pathogen-evoked responses.

We agree and have performed complementary experiments showing induction of CCRICs by EPEC and eATP in polarized intestinal epithelial cells, now shown in Figure S8.

Version 2

Abstract

Enteropathogenic Escherichia coli (EPEC) is a major bacterial enteropathogen causing infectious diarrhea among children in developing countries. Here, we found that EPEC induced isolated Ca2+ responses in epithelial cells, triggered by extracellular ATP (eATP). These responses were dependent on type III secretion (T3S) and down-regulated by the bacterial secreted protease EspC, consistent with eATP released by the T3S translocon pore-forming activity in host membranes. By performing high speed Ca2+ imaging, we uncovered that at the onset of infection, low eATP levels triggered Ca2+-responses involving the whole cell but showing the small amplitude and fast kinetics usually associated with local Ca2+ responses. The findings, supported by theoretical modeling, evocate a conceptual shift whereby low amounts of inositol 1, 4, 5-trisphosphate (IP3) induced by low eATP levels and subsequent moderate Ca2+ release enable the fast coordination of IP3 receptor cluster activation throughout the cell. Importantly, these yet undescribed coordinated fast responses occurred over prolonged time periods and defined a cell state with dampened activation of the pro-inflammatory transcriptional activator NF-kB associated with a decrease in its Ca2+-dependent O-linked β-N-acetylglucosamine modification.

Introduction

EPEC are diarrheagenic E. coli strains that cause significant morbidity and mortality in children under two years of age. While infection rates have significantly declined in industrialized nations, EPEC remains a major public health concern in low-income countries (Lozer et al., 2013). EPEC form attaching and effacing (A/E) lesions on intestinal epithelial cells and lack the ability to produce Shiga toxins or heat-labile (LT) and heat-stable (ST) enterotoxins (Croxen et al., 2013; Gomes et al., 2016; Hazen et al., 2016). The ability of EPEC to form A/E lesions is determined by the locus of Enterocyte Effacement (LEE), a large genomic pathogenicity island that encodes the essential genetic elements required for this process (Pearson et al., 2016). The LEE region of EPEC (E2348/69) encodes components of the type III secretion system (T3SS), a molecular apparatus that translocate at least 25 bacterial effector proteins into the host cell. The EPEC type III secretion system (T3SS) consists of a basal body and a needle-like structure, resembling those found in Salmonella and Shigella, but with a distinct sheath-like extension at the needle tip, primarily composed of EspA, and about ten times longer than the T3SS needles from other bacterial species. This EspA filament acts as a molecular bridge, extending from the bacterium to the host cell membrane, allowing insertion of the EspB and EspD translocon components into the host cell membrane, enabling type III effectors injection in the cell cytosol (Creasey et al., 2003; Monjarás Feria et al., 2012; Sal-Man et al., 2012). Osmoprotection assays suggest that the translocon forms a pore with an internal diameter ranging between 3 to 5 nm, allowing the passage of unfolded effector proteins into the host cell (Chatterjee et al., 2015). However, the T3SS translocon shows weak pore-forming activity during EPEC infection of epithelial cells, presumably because it forms a sealed conduct between the T3SS and host cell membranes (Guignot et al., 2016). EspC, a secreted serine protease from the autotransporter family, targets EspA and EspD and down-regulates pore formation activity associated with cytotoxicity (Guignot et al., 2015). EPEC T3SS effector proteins translocated into host cells lead are responsible for attaching and effacing (A/E) lesions and intimate bacterial adhesion to the host cells associated with the formation of an actin-rich pedestal formation (Chen & Frankel, 2005).

The detection of pathogenic bacteria by intestinal epithelial cells plays an important role in initiating pro-inflammatory responses. Recognition of bacterial surface components by pattern recognition receptors triggers pro-inflammatory signaling pathways involving the transcriptional activator NF-κB and the production of cytokines such as interleukin-8 and tumor necrosis factor-α (TNF-α) (Edwards et al., 2011). However, EPEC suppresses these signaling pathways early in infection through the coordinated action of several T3SS effector proteins. Among the first translocated T3SS effectors, Tir interacts with TNF-α receptor-associated factors (TRAF2 and TRAF6), recruiting the tyrosine phosphatases SHP-1 and SHP-2 (Mills et al., 2008; Ruchaud-Sparagano et al., 2011; Yan et al., 2013). Several non-LEE T3SS effectors, including NleE, NleB, NleH1, NleH2, NleC, and NleD, further contribute to inhibiting NF-κB and MAPK signaling. NleE and NleB stabilize the interaction between NFκB and its inhibitory subunit IκB, preventing its degradation, thereby keeping NFκB in an inactive state. The ability of NleE to inhibit NFκB signaling depends on its S-adenosyl-L-methionine (SAM)-dependent methyltransferase activity (Zhang et al., 2011). NleE-mediated methylation of TAB2/3 prevents IKK activation (Zhang et al., 2011). NleB selectively blocks NFκB activation through GlcNAcylation of the TNF-α receptor (TNFR) adaptor protein and TNFR1-associated death domain (TRADD) (S. Li et al., 2013; Pearson et al., 2013). The T3SS effectors NleH1, NleH2 and NleC also interfere with NFκB nuclear translocation (Gao et al., 2009). NleC specifically cleaves P65 RelA (Giogha et al., 2015; Ruchaud-Sparagano et al., 2011; Yen et al., 2010). Interestingly, NleF has been implicated in the activation of NF-κB, underscoring the complexity of the regulation of inflammation during bacterial infection and suggesting the timing of various T3SS effectors’ activity (Pallett et al., 2014).

A similar complexity applies to T3SS effectors regulating cell death and survival pathways during EPEC infection of epithelial cells. Tir was found to elicit a rapid Ca²⁺ influx across the host cell membrane, through the activation of a host plasma membrane Ca²⁺ channel, the mechanosensitive transient receptor potential vanilloid 2 (TRPV2) leading to pyroptosis (Zhong et al., 2022). However, the NleA effector, blocks the delivery of TRPV2 channels to the cell surface, thereby dampening Tir-induced Ca²⁺ influx. The effector NleF also directly binds caspase-4 to inhibit its activity (Zhong et al., 2020). The extrinsic apoptotic pathway is triggered by EPEC pili (Abul-Milh et al., 2001). However, the NleD and NleB effectors inhibit this pathway by cleaving JNK and GlcNAcylating the death domain adaptor proteins TRADD and FADD, respectively (Baruch, Gur-Arie, et al., 2011; Pearson et al., 2013). While EspC prevents cytotoxicity linked to pore-formation by the T3SS translocon during the early EPEC infection phases, it was shown to promote intrinsic apoptosis through increase in intracellular Ca2+ and calpain activation (Serapio-Palacios & Navarro-Garcia, 2016).

Central to inflammation and cell death / survival pathways induced by EPEC, bacterial-induced Ca2+ signals have been a matter of debate. EPEC infection is known to perturb host Ca²⁺ signaling, but the source and sequence of Ca²⁺ signals during infection remain controversial. EPEC was shown to induce Ca²⁺ influx associated with a loss of mitochondrial membranes permeability leading to cell death (Zhong et al., 2020; Ramachandran et al., 2020; Zhong et al., 2022), but was also reported to trigger IP3-mediated Ca²⁺ release possibly involved in bacterial-induced cytoskeletal rearrangements (Baldwin et al., 1991; Baldwin et al., 1993; Foubister et al., 1994; Bain et al., 1998).

Here, we investigated the characteristics and implications of EPEC-induced Ca²⁺ responses in epithelial cells. We characterized yet undescribed Ca²⁺ signals induced by EPEC and low ATP levels, presenting the fast dynamics and small amplitude of local Ca²⁺ responses but involving large cell area. We found that these responses likely result from the coordination of elementary responses via rapid Ca2+-induced Ca²⁺ release over large cell area, challenging generally admitted concepts on Ca²⁺ diffusion. Importantly, we show that these newly described responses have functional implications by dampening the cell ability to respond to inflammatory signals.

Results

EPEC induces Ca2+ responses that depend on Type III secretion-mediated eATP release

Despite their critical role in cellular processes key to bacterial infection, EPEC-induced Ca²⁺ responses remain to be characterized. We therefore set up to perform a detailed single cell imaging of Ca²⁺ responses elicited by cells infected by EPEC.

As shown in Fig. 1, EPEC induced Ca²⁺ transients often corresponding to a single peak of varying amplitude detected over several minutes, corresponding to 6.2 ± 0.8% (mean ± SEM) of the maximal histamine response (Figs. 1A, B). These Ca²⁺ responses were dependent on a functional T3SS, since they were not observed for the T3SS-deficient escN mutant (Figs. 1A, C). Only 40 ± 4.7 % (mean ± SEM) of cells, however, elicited responses when challenged wild-type EPEC at a low multiplicity of infection (MOI) of 20 bacteria per cell, a value that raised to 76 ± 4.8 % (mean ± SEM) when using a high MOI of 80 bacteria per cell (Fig. 1C). In contrast, even at the low MOI, more than 83 % of cells showed actin pedestals, indicating that Ca²⁺ responses were elicited only in a fraction of cells targeted by EPEC-type T3SS (Figs. 1D, E). The frequency of Ca²⁺ responses per cell increased over the incubation time with an average frequency of responses per cell raising by 6.9 to 8.3-fold from the first to the last 30 min of EPEC challenge (Fig. 1F). This increased frequency suggested the accumulation of an agonist in the extracellular medium during the course of the infection triggering IP3-mediated Ca²⁺ release. When pooling all single cell responses, we could observe a steady increase in the average cytosolic Ca2+ concentration of the cell population, as previously reported (Ramachandran et al., 2020; Figs. S1B). However, when performing single cell imaging, we did not observe an increase in cytosolic Ca²⁺ basal levels even at high MOI after 2 hours incubation with wild-type bacteria (Fig. S1A).

EPEC induces isolated Ca2+ responses of limited amplitude in epithelial cells.

HeLa cells were loaded with the fluorescent indicator Cal-520, challenged with the indicated bacteria and subjected to live-cell Ca2+ imaging at a frequency of one acquisition every 10 seconds (A-C, F) or fixed and processed for fluorescence microscopy analysis (D-E) (Materials and Methods). A, Representative traces of Ca2+ variations in single cells. The black arrowheads indicate the time of bacterial challenge. The blue arrowheads indicate stimulation with 3 µM histamine. B, Response amplitude expressed as a percent of the maximal histamine response amplitude (N = 3, n > 63). C, Percent of cells exhibiting Ca2+ responses (N =3, cells > 66). (D, E) Cells challenged with RFP-expressing bacteria for 1 hour. D, Representative confocal micrographs. Staining with DAPI (blue), phalloidin-Alexa 488 (green). The lower panels show a higher magnification of the insets in the top panels. Scale bar = 10 µm. E, Percentage of bacteria-associated actin-rich pedestals (N = 3, cells > 273). F, average number of responses per cell during the first 30 min (0-30) and last 30 min (30-60) of bacterial challenge. Low MOI: 10 bacteria / cell. High MOI: 50 bacteria / cell. Bar: mean. (N = 3, cells > 63). Mann-Whitney test. *: p < 0.05; **: p < 0.01; ***: p < 0.001; ****: p < 0.0001.

Together, these results suggest that EPEC induces isolated Ca²⁺ responses that depend on the T3SS for the release of limiting amounts of a Ca²⁺ agonist.

EPEC-mediated Ca²⁺ responses depend on ATP released in the extracellular medium via the T3SS translocon

In previous works, we showed that the EPEC T3SS translocon forms pores in host cell plasma membranes that were down-regulated by the bacterial secreted serine protease EspC (Guignot et al., 2016). We posit that low amounts of ATP released in the extracellular medium by the T3SS translocon were responsible for the isolated Ca²⁺ responses of reduced amplitude elicited by EPEC. According to this view, by removing T3SS translocons from host cell membranes, EspC would down-regulate EPEC-mediated Ca2+ signaling explaining the low ratio of Ca2+ responding cells relative to cells forming actin pedestals.

Consistent with this and as shown in Figs. 2A-C, an espC mutant induced more Ca²⁺ responses than wild-type EPEC, with 94 ± 3% responding cells and a frequency of 10.5 ± 1.2 responses per cell over the 60 min analysis, compared 40 ± 4.7% responding cells and less than 2 responses per cell for the espC mutant and wild-type EPEC, respectively. Also, the average amplitude of Ca²⁺ responses induced by the espC mutant was higher than that of wild-type EPEC, suggesting more eATP release (Fig. 1B). Accordingly, cell treatment with the purinergic receptors’s antagonists Suramin and PPADS as well as with hexokinase to deplete eATP, inhibited Ca²⁺ responses induced by the wild-type and espC mutant strains (Figs. 2B, C, S2A-C). Treatment with the PLC inhibitor U73122, but not its inactive analog U73343, also resulted in inhibition of EPEC-mediated Ca²⁺ responses (Figs. S2A, B). As expected for ATP-mediated Ca2+ release, sample treatment with EGTA, a cell impermeant chelator of extracellular Ca2+ did not decrease the percent of Ca2+ responding cells triggered by wild-type EPEC or the espC mutant (Fig. S2C). The frequency of responses per cell was also not inhibited and even appear to increase upon EGTA-treatment in cells challenged with wild-type EPEC and the espC mutant (Figs. S2D, E). In control experiments, Suramin treatment did not affect actin pedestal structures induced by these strains (Fig. S3).

EPEC-induced Ca2+ responses are elicited by ATP released by the T3SS translocon

HeLa cells were loaded with the fluorescent indicator Cal-520 or with 200 μM suramin for 30 minutes, challenged with the indicated bacteria and subjected to live-cell Ca2+ imaging for a 60 min-duration at a frequency of one acquisition every 10 seconds. A, Representative traces of Ca2+ variations in single cells. The black arrowheads indicate the time of bacterial challenge. The blue arrowheads indicate stimulation with 3 µM histamine. B, Percent of cells exhibiting Ca2+ responses (N = 3, cells > 70). C, Average number of responses per cell. Low MOI: 10 bacteria / cell. High MOI: 50 bacteria / cell. Bar: mean. (N = 3, cells > 30). Mann-Whitney test. **: p < 0.01; ***: p < 0.001; ****: p < 0.0001.

Together, these results suggest that Ca²⁺ responses induced by EPEC are mediated by ATP released in the extracellular medium via pores formed by the T3SS translocon and are down-regulated by EspC.

EPEC induces coordinated Ca²⁺ responses from single IP3R clusters

We previously showed that Shigella induced local Ca2+ responses dependent on the T3SS and Ca2+ release (Tran Van Nhieu et al., 2013), suggesting that insertion of the Type III translocon was responsible for bacterial-induced local Ca2+ signals. We therefore set up to investigate whether EPEC could also trigger T3SS-dependent local Ca2+ responses.

To explore this, we performed rapid Ca2+ imaging at a frequency of 57 ms acquisition per frame to sample elementary Ca2+ release events. As shown in Fig. 3, by performing high speed Ca2+ imaging, we detected fast Ca2+ increases associated with cell challenge with EPEC. These fast Ca2+ responses did not correspond to other responses previously reported since they involved the whole cell or a large cell area but showed a small amplitude and fast dynamics usually associated with local Ca2+ responses (Fig. 3B). As observed for the ATP-dependent responses shown in Fig. 2, the espC mutant triggered a higher percent of Ca2+ responding cells than wild-type EPEC (Fig. 3C). Also, the response amplitude was higher for the espC mutant with an average amplitude corresponding to 7.7 ± 0.4 % (mean ± SEM) of the maximal agonist response, compared to 5.4 ± 0.4 % (mean ± SEM) for wild-type EPEC (Fig. 3E). These responses occurred repeatedly at a high frequency of up to 4.5 responses per minute during several minutes following bacterial challenge (Figs. 3D and 3F) and were dependent on Type III Secretion, as evidenced by the lack of response in cells infected with the ΔescN strain (Figs. 3B and 3C). EPEC-induced fast Ca2+ responses were dependent on Ca2+ release since they were inhibited by U73122, a PLC inhibitor (Figs. 3C and 3D). These similarities with the EPEC-induced eATP-dependent Ca2+ responses suggested that the atypical fast responses were triggered by low amounts of ATP released in the extracellular medium following insertion in host cell plasma membranes of discrete numbers of EPEC T3SS translocons. Consistently, these atypical EPEC-induced fast responses were abolished in the presence of the ATP receptor inhibitor Suramin (Figs. 3C, D).

EPEC induces rapid and Coordinated Elementary Ca2+ Responses.

HeLa cells were loaded with the fluorescent indicator Cal-520, challenged with the indicated bacteria and subjected to high speed Ca2+ imaging at a frequency of one acquisition every 57 ms for a duration of 110 seconds. A, Representative time-serie of pseudocolored fluorescent micrographs of cells challenged with wild-type EPEC. The numbers indicate the elapsed time in ms from an arbitrarily determined origin. Scale bar = 10 µm. B, D, traces of Ca2+ variations in 2 subcellular regions of the same cell. B, WT: traces corresponding to the regions depicted in the image 0 of panel A. The arrowheads point to the Ca2+ responses shown in Panel A with the corresponding color. C, Percent of cells exhibiting Ca2+ responses (N > 3, cells > 134). + Suramin: treatment with 200 μM Suramin. +U73122: treatment with 10 μM U73122. E, Response amplitude expressed as a percent of the maximal response amplitude induced by treatment with 3 μM histamine (N = 3, cells > 166). F, average number of responses per cell. C, E, Bar: mean. Mann-Whitney test. **: p < 0.01; ****: p < 0.0001. F, High speed Ca2+ imaging was performed every 5 min for 110 seconds following infection with the indicated bacterial strain as depicted the scheme. The average number of responses per cell is indicated (N > 3, cells > 29).

Together, these results show that at the onset of infection, EPEC induces an atypical pattern of fast Ca2+ responses involving the whole or a large area of the cell, likely resulting from low ATP levels released by the insertion of a discrete number of translocons in host cell membranes.

EPEC-induced fast Ca2+ responses are triggered by low ATP levels

Previous studies have described local Ca²⁺ increases triggered by sub-maximal agonist concentrations and leading to limited IP3-mediated Ca2+ release. These local Ca²⁺ responses are typically small, of short durations and localized to subcellular regions. Among these, the so-called “Blips” correspond to elementary events of opening of a single IP3 receptor channel usually lasting between 50 and 100 ms, whereas “Puffs” involve the synchronized activation of multiple IP3 receptor channels in localized clusters and last several hundreds of ms (Swillens et al., 1999). In contrast to these described local Ca2+ signals, EPEC-induced fast and small responses could occur throughout the cell, suggesting the coordination of elementary responses over large cell area. Since our findings suggested that these responses were elicited by low amounts of eATP released by a discrete number of T3SS translocons, we investigated whether low ATP levels could elicit similar responses.

HeLa cells treated with 150 nM ATP showed Ca2+ responses that were indistinguishable from fast responses elicited by EPEC, with an average percent of Ca2+ responding cells of 61.2 ± 5.8 % (mean ± SEM) and a frequency of 3.9 responses per cell over 60 seconds (Figs. 4A, C and S4A). These fast Ca2+ responses had an amplitude that did not exceed 10 % of the maximal agonist response and occurred over several minutes (Fig. 4A), with a duration of 2.1 ± 1.0 sec (mean ± SEM) (N = 4, 128 responses). As observed for EPEC, fast Ca2+ responses induced by low ATP levels involved the whole cell or large cell area encompassing the nuclear and perinuclear area and corresponding to at least 30 % of the cell area as illustrated in Fig. 4B. In this large area, all ROIs corresponding to 1 square micron showed a superimposable profile, as illustrated by traces in Fig. 4C. Similar fast and small coordinated Ca2+ responses were also observed when cells were challenged with 100 nM of histamine, another Ca2+ agonist, with 44 ± 7 % (mean ± SEM) of cells showing responses, suggesting that these are generic responses triggered by low levels of IP3 (N = 2, cells = 340; Fig. S4). EPEC and low concentrations of eATP induced similar responses in polarized intestinal epithelial cells (Fig. S5).

EPEC-induced Coordinated Elementary Ca2+ Responses are reproduced by low ATP levels.

A-C, HeLa cells were loaded with the fluorescent indicator Cal-520, challenged with 150 nM ATP and subjected to Ca2+ imaging. Image acquisition every 52 ms (A) or 22 ms (B, C). A, Traces of Ca2+ variations corresponding to a single cell (red trace), or subcellular regions within the same cell (inset). A, arrowhead: challenge with 2 μM ATP. B, Time serie of fluorescent micrographs pseudocolored using the “glow” Fiji lookup table, where the blue pixel correspond to an arbitrarily set threshold value. The numbers indicate the elapsed time in seconds. Blue: high intensity pixels showing the large top cell area with CCRICs and the local lower puff area. Scale bar = 10 µm. C, Traces corresponding to Ca2+ variations in the subcellular regions depicted in Panel B. The responses are labelled 1-4, with the response 1 corresponding to the puff (Panel B, red ROI) impulsing the response 3 in the same region. Responses 2 and 4 correspond to CCRICs in Panel B, blue and green ROIs. Note the diffusion of the responses from the initial release area in other area inferred from the dampening of the response amplitude.

More detailed scrutinizing showed that in their initial mounting phase, these fast Ca2+ responses were created by the opening of discrete clusters involving an area of ca. 0.04 μm2 that had a transient activity, or possibly were highly mobile, since they were seldom detected at a similar location for three consecutive 22 ms acquisition frames (Figs. S6A, B). These discrete clusters showed similar Ca2+ kinetics suggesting the coordination of Ca2+ release of single IP3R clusters throughout the area that we will hereafter termed CCRICs for “Coordinated Ca2+ Responses from IP3R Clusters”. Treatment with BAPTA-AM to chelate intracellular Ca2+ led to a complete inhibition of CCRICs (N = 3, n > 150 cells; Fig. S7A). Ca2+ responses could still be detected upon cell treatment with EGTA-AM consistent with its lower kon rate for Ca2+, but with a significant inhibition of the percentage of Ca2+ responding cells as well as of the frequency of responses per cell, suggesting that coordination could occur via Ca2+ diffusion and Ca2+-induced Ca2+ release (Fig. S7).

In rare instances (less than 3%), typical local “Puff” responses elicited by these ATP concentrations could also be detected often occurring at the cell periphery (Figs. 4B, red region and 4C, red arrow; Fig. S6D, blue trace) (N > 20, cells > 500). As expected from the small concentrations of Ca2+ released at puff sites, no increase in cytosolic Ca2+ was detected in a distal cell region (Fig. S6D, top), indicating that isotropic Ca2+ diffusion from a puff release site cannot account for Ca2+ increase over large cell area. Puffs could also be detected concomitantly with CCRICs in different ROIs of the same cell (Fig. S6D, bottom). In contrast to puffs, CCRICs often showed responses of comparable amplitude in distal regions over the whole cell (Figs. 4C and S6A, B), suggesting the contribution from IP3R cluster activation by Ca2+-Induced Ca2+ Release (CICR). Within a given cell, the vast majority of CCRICs appeared quasi-synchronized at the fatest acquisition rate of 22 ms / frame that we could achieve. However, in few instances a delay could be detected in the elicitation of a peak in distant region of a cell (Fig. S6C). These observations suggest that the quasi-synchronization of CCRICs result from the fast diffusion of Ca2+ leading to the activation of IP3R clusters over large cell area, which may be delayed in a some instances. Scrutinizing of CCRICs showed that while their profiles were comparable, the amplitude of these responses varied in different regions of the cell, with often a single 1 μm2 region, likely corresponding the initial firing cluster, showing a prominent amplitude and other regions with smaller amplitude for a given response (Figs. 4B and 4C). For example, in Fig. 4C, the highest amplitude is observed in the red region for peaks 1 and 3, whereas it is observed and in the purple region for peak 2. Thus, for a given CCRIC, the respective contribution of local IP3R cluster activation and isotropic diffusion of Ca2+from other release sites in Ca2+ increase may vary in different regions of the cell.

CCRICs are coordinated by the rapid diffusion of Ca2+ at low concentrations in cell area with a high density of IP3 clusters

In Fig. 5, we used modeling to further investigate the mechanism of coordination of these fast Ca2+ responses. Based on our previous studies (Voorsluijs et al., 2019; Ornelas-Guevara et al., 2023), the model provides a fully stochastic spatial description of Ca2+ release dynamics from IP3R clusters in a two-dimensional representation of a HeLa cell. The simulation domain extends on 10 × 10 µm2 and is discretized into a 20 × 20 grid of compartments (0.5 × 0.5 µm2 each), each representing a cytosolic subvolume of 10-16 L. Each compartment contains at most one cluster of IP3R, whose dynamics is described as a whole. Besides, cytosolic Ca2+ concentration can also vary because of uptake by SERCA, release by a leak or diffusion (Supplementary Information, Note 1).

Modeling of Coordinated Elementary Ca2+ Responses.

Top, Ca2+ variations in subcellular area within a single cell are represented in pseudocolor. Shown are the maximum values of Δ[Ca2+]/[Ca2+]b reached in each compartment during a 60s simulation. Empty white squares: IP3R clusters. Graphs, Traces correspond to Ca2+ variations in the region with the matching color. Colored arrows: Ca2+ response due to the activation of an IP3 cluster in the region with the matching color. Colored arrowhead and dashed red and blue lanes: Ca2+ variations due to the diffusion of a Ca2+ response from or nearby to the region with the matching color. Black arrows: Ca2+ response due Ca2+-activated Ca2+ release. A, low density of IP3R clusters with local responses detected. B, C, Empty black box: area with a high density IP3R clusters. C, similar to B, but following IP3R cluster sensitization due to increased Ca2+ responses.

In Fig. 5A, low ATP levels lead to low IP3 levels activating a limited number of IP3 clusters, opening stochastically and releasing small amounts of Ca2+. In a given area of the cell, the Ca2+ variation integrates Ca2+ release from clusters within this area, as well as Ca2+ diffusing from or to other cell area. For a given response, the initially firing cluster is contained in an area characterized by the highest Ca2+ peak amplitude (Fig. 5A, blue and green arrows) that dampens in a distal area (Fig. 5A, blue and green arrowheads). If the density of IP3R clusters is low, as expected for ER compartment at the cell periphery, the spatial segregation of the initial firing cluster and resulting Ca2+ diffusion to other area is clearly detected (Fig. 5A). In instances, however, the model predicts temporally coordinated responses of similar amplitude, suggesting Ca2+-induced Ca2+ release from secondary clusters (Fig. 5A, black arrow). For both type of Ca2+ dynamics, a large value of the Ca2+ diffusion coefficient of 100 μm2 / s is a key parameter that needs to be taken into account in the model. While it is generally admitted that Ca2+ diffuses very slowly due to the Ca2+ buffers in the cell (∼30 μm2/s), the low levels of Ca2+ released by CCRICs may not be subjected to the diffusion limitations observed at higher Ca2+ levels, because of the relative moderate affinity of buffers for Ca2+. How the effective diffusion coefficient of Ca2+ is affected by the Ca2+ concentration is explained in more detail in Supplementary Information, Note 2.

If the IP3R cluster density is high, as expected in the large perinuclear and nuclear area corresponding to the bulk of the ER, the coordination between individual clusters is very fast (Fig. 5B). As a result, the identification of initial firing clusters goes beyond the technical capacities of the imaging set-up, and ROI within this area show comparable profiles of fast Ca2+ responses (Fig. 5B). Upon prolonged incubation with increasing Ca2+ responses and sensitization of IP3R clusters, the coordination of responses linked to Ca2+-induced Ca2+ release becomes predominant throughout the cell (Fig. 5C). Moreover, at high IP3 concentrations, the model reproduces the propagation of a large-amplitude Ca2+ wave, as expected (Supplementary Information, Note 3).

Low eATP levels dampen NF-κB activation

Our findings indicate that the novel Ca²⁺ response pattern is not exclusive to EPEC infection and can be replicated by low levels of eATP or histamine, suggesting a broader physiological relevance. From an immunological perspective, CCRICs may therefore play a critical role in various signaling pathways during bacterial infection. eATP is a well-characterized danger signal contributing to the elicitation of pro-inflammatory signals in various tissues in response to infections (Savio et al., 2018). Previous studies linked intracellular Ca²⁺ signaling and NF-κB activation, a key transcription factor that triggers inflammatory responses (Smedler et al., 2014). We therefore set up to investigate the effects of CCRICs triggered by low eATP levels on NF-κB activation, by performing Western blot analysis against the phosphorylated forms of IκBα (p- IκBα) and P65 (p-P65).

As shown in Figs. 6A and 6B, in control samples, TNF-α induced IκB-α phosphorylation peaking 10 minutes following challenge. In contrast, in the presence of low ATP levels, IκB-α showed a delayed phosphorylation with a 2.1 fold decrease at 10 minutes post-challenge (Figs. 6A and 6B). Consistently, the rates of IκB-α degradation were also slower in the presence of ATP relative to control (Figs. 6A and 6C). As expected from the IκB-α results, TNF-α induced the phosphorylation of the NF-κB P65 subunit and ATP led to a delay and decrease in P65 phosphorylation (Figs. 6D and 6E).

Low ATP levels dampen NF-kappaB activation.

HeLa cells were stimulated with 10 ng/ml TNF-α alone or in the presence of 150 nM ATP or 20 μM BAPTA-AM (F-H). At the indicated time points, cell lysates were analyzed by Western blot using the indicated antibodies. A, D, F, G, Representative blots. B, C, E, H, Densitometry analysis of the indicated antibody signal normalized to that of HSP90 (B, C) or total P65 (E, H), expressed as fold-increase to basal levels of p-IκB (B), IκB (C) or p-p65 (E, H) at time = 0. Values correspond to the mean ± SEM of 3 or 4 independent experiments. p-p65: anti-phospho P65 antibody. p-IκB: anti-phospho IκB antibody. ANCOVA test. *: p < 0.05; **: p < 0.01; ***: p < 0.001.

In control experiments, we did not detect differences in P65 phosphorylation in response to TNF-α stimulation when cells were treated with BAPTA-AM to chelate intracellular Ca2+ (Figs. 6F and 6H). However, cell treatment with BAPTA-AM prevented the dampening of P65 phosphorylation triggered by low ATP levels (Figs. 6G and 6H), suggesting that CCRICs down-regulated TNF-α -induced NF-κB activation.

Low eATP levels down-regulates NF-κB activation through Ca2+-dependent O-GlcAcylation

We next investigated how CCRICs could regulate NF-κB activation. O-linked β-N-acetylglucosamine (O-GlcNAc) transferase (OGT) was reported to regulate NF-κB signaling by post-translationally modifying the p65 subunit (Ruan et al., 2017). Interestingly, OGT is regulated by Ca2+ signaling, suggesting that CCRICs could affect NF-κB activation via O-GlcNAcylation.

As shown in Fig. S6, TNF-α in the presence of 150 nM eATP stimulated the levels of O-GlcNacylation, specifically for proteins with an apparent molecular weight superior to 100 kDa that was not observed with TNF-α alone. To further investigate the effects of low eATP levels on NF-κB O-GlcNacylation, we performed immunoprecipitation of P65 RelA on lysates of cells stimulated for 12 minutes with TNF-α alone or co-stimulated with TNF-α and 150 nM eATP. As shown in Fig. 7, TNF-α induced increased O-GlcNacylation of P65 relative to non-stimulated cells, but this increase was inhibited by low eATP levels. Inhibition of P65 O-GlcNacylation by eATP was Ca2+ dependent, since it was not observed in the presence of BAPTA-AM (Fig. 7).

Low ATP levels down-regulate NF-kB O-GlcNAcylation in a Ca2+-dependent manner.

HeLa cells were stimulated with 10 ng/ml TNF-α alone or in the presence of 150 nM ATP with or without 20 μM BAPTA-AM for 12 min. Cell lysates were subjected to P65 immunoprecipitation. A, Representative blots with the indicated antibodies. IP: immunoprecipitates; L: total cell lysates. B, Densitometry analysis of the O-GlucNAc signal in P65 immunoprecipitates normalized to that of TNF-α alone. Mann-Whitney test. N = 4. *: p < 0.05. ns: not significant.

The results indicate that low eATP levels inhibit O-GlcNacylation of P65 induced by TNF-α in a Ca2+-dependent manner and suggest that differential O-GlcNacylation of P65 relative to higher molecular weight proteins.

Discussion

We report here the characterization of CCRICs, corresponding to yet undescribed Ca2+ responses. CCRICs showed rapid kinetics with an average duration of ca 2.1 seconds and amplitude corresponding to an increase in Ca2+ cytosolic concentration of a few hundreds nM, seemingly smaller than that of puffs (Fig. S6D), often occurring repeatedly with a frequency of up to 12 CCRICs / min over the whole cell. Our modelling studies support the notion that CCRICs implicate the rapid coordination of IP3R clusters via CICR in large cell area, challenging established concepts in the Ca2+ signaling field.

Ca²⁺ diffusion in the cytosol is regulated by mobile and immobile Ca²⁺-binding proteins acting as buffers and forming localized microdomains with steep Ca²⁺ concentration gradients. At the mouth of an open IP₃R channel, Ca²⁺ concentration can reach 100 μM, while just 1–2 μm away, it may drop below 1 μM. Diffusion is restricted by Ca2+ buffers with KD’s that are generally lower than this concentration. Consequently, Ca²⁺ signaling is generally admitted to be spatially restricted, typically influencing regions within approximately 5 μm of the release site (Foskett et al., 2007). The distribution of Ca²⁺-binding proteins and the spatial arrangement of release channels allow IP₃R-mediated [Ca²⁺]i signals to exhibit diverse spatial and temporal properties, making this system highly adaptable (Vandeput et al., 2007). High-resolution optical imaging of fluorescent Ca²⁺ indicators in intact cells indicates that IP₃-mediated [Ca²⁺]i signals are structured levels (Foskett et al., 2007). At low IP₃ levels, individual IP₃R opens stochastically at discrete release sites, causing localized elevations in cytoplasmic [Ca²⁺]. At higher IP₃ levels, Ca²⁺ release spreads between IP₃R clusters, propagating waves that travel at tens of microns per second, coordinating intracellular signaling ultimately leading to a global Ca²⁺ response. In reference models describing intracellular Ca2+ dynamics, cell regions with a high IP3R density initiate the Ca2+ response from which Ca2+ waves propagate with a diffusion coefficient of 10 – 30 μm2/s (Falcke et al., 2003).

In contrast to these described global and local Ca²⁺ responses, we found CCRICs to be highly temporally coordinated over large area, suggesting the fast diffusion of Ca2+ and propagation of Ca2+ by CICR. While challenging generally admitted concepts on the poor diffusion of Ca2+, this view is fully supported by our theoretical modeling implicating the fast diffusion of Ca2+, with a diffusion coefficient of at least 100 μm2/s that can be expected at low cytoplasmic [Ca²⁺]. Indeed, at these low Ca²⁺ concentrations not exceeding a few hundreds nM, the majority of Ca²⁺ buffers are not expected to efficiently bind to Ca²⁺ and to significantly interfere with Ca²⁺ diffusion because of their relative low affinity.

We found that CCRICs implicate a large cell area including the nuclear and perinuclear area. The large CCRIC area likely involves the bulk of the endoplasmic reticulum, while peripheral area may contain smaller ER compartments. In our model considering fast Ca2+ diffusion when buffers are far from saturation, the higher density of IP3R clusters in the CCRIC area accounts for the high coordination of the responses, relative to the lack of coordination in peripheral area where lower IP3R density is expected. Consistently, while CRICCs were detected in the vast majority of cells at these very low agonist concentrations, in rare instances, local “puff-like” responses were also detected at the cell periphery. These observations are in contrast to previously described Ca2+ puffs preceding global responses reported to occur preferentially in perinuclear area (Thomas et aL., 1999). These earlier studies, however, involved higher agonist concentrations (1-5 μM ATP) expected to lead to the release of higher IP3 concentrations, which may preferentially stimulate larger IP3R clusters at the perinuclear region because of the higher density of IP3Rs. In addition, larger IP3 clusters may release higher amounts of Ca2+ for which, as opposed to CCRICs, diffusion would be restrained by Ca2+ buffers thereby favoring the spatial confinement of the response.

We showed that a low dose of eATP triggering CCRICs delayed and dampened NF-κB activation linked to a reduction in O-GlcNAcylation of the NF-κB p65 subunit. One major open question is the mechanism by which CCRICS down-regulate NF-κB activation. NF-κB activity can be modulated by O-GlcNAcylation, a reversible glycosylation modification catalyzed by O-GlcNAc transferase (OGT) and O-GlcNAcase (OGA) (Liu & Ramakrishnan, 2021). Previous studies demonstrated that Ca²⁺ signals activate Ca²⁺-regulated enzymes like CaMKII, which in turn phosphorylates and activate OGT, promotes O-GlcNAcylation (Ruan et al., 2017). In other studies, OGT-mediated O-GlcNAcylation could modulate NF-κB signaling pathway (X. Dong et al., 2023). O-GlcNAcylation of P65 could also inhibit its interaction with IκB-α, promote p65 nuclear translocation and increase NF-κB transcriptional activity (Liu & Ramakrishnan, 2021). Reduced O-GlcNAcylation of P65 at residues S550 and S551 was shown to result in decreased NF-κB activation and nuclear translocation (Motolani et al., 2023). These findings are in line with our results suggesting that CCRICs elicited by low-level eATP, downregulate p65 O-GlcNAcylation and NF-κB activation, possibly by modulating OGT activity or OGT-p65 interactions. Reduced O-GlcNAcylation of p65 may affect its phosphorylation patterns indirectly, perhaps by altering the interaction of NF-κB with kinases or phosphatases involved in its activation (Özcan et al., 2010).

Our findings indicate that eATP differentially regulate inflammatory signaling pathways in epithelial cells by dampening NF-κB activation at low levels and stimulating its activation at high concentrations. These results extend the key role of eATP from a danger-associated molecular pattern (DAMP) to a fine-tuner of inflammatory responses depending on its concentration.

Materials and Methods

Cell and Bacterial culture

HeLa cells were maintained in Dulbecco’s Modified Eagle Medium (DMEM; Gibco, Thermo Fisher Scientific) supplemented with 10% fetal bovine serum (FBS; Gibco, Thermo Fisher Scientific) Human colon adenocarcinoma Caco-2/TC-7 cells were maintained in DMEM containing 20 % FBS, supplemented with non-essential amino acids. Cells were grown at 37 °C in a humidified incubator with 10% CO₂. Wild-type Enteropathogenic Escherichia coli (EPEC WT), ΔescN, and ΔespC strains were cultured in Luria-Bertani (LB) broth at 37 °C with kanamycin at a final concentration of 15 μg/ml in a shaking incubator. All strains were transformed with the pmCherry-N1 plasmid, which encodes red fluorescent protein and carries an ampicillin resistance gene for selection. Where applicable, ampicillin (100 μg/mL) and kanamycin were included to maintain resistance markers. The red fluorescence enabled visualization of the bacteria during downstream analyses.

EPEC infection of cells

HeLa cells were seeded in 6-well plates at a density of 4.5 × 10⁵ cells per well one day before infection. TC-7 cells were seeded at a density of 5 × 10⁵ cells / well in 6-well plates and allow to polarize for 4 days prior to bacterial challenge, replacing medium every day. EPEC WT, ΔescN, and ΔespC strains grown in the exponential phase were resuspended and primed for 5 hours before challenging HeLa cells in DMEM medium. Cells were challenged with bacteria at an OD600 = 0.2 (Low MOI) or 0.8 (High MOI).

Ca2+ imaging

HeLa cells were seeded onto 25 mm-diameter glass coverslips. Cells were preloaded with the fluorescent indicator dye Cal-520 (AAT #21130) for 30 minutes at room temperature, followed by two PBS washes and one time with DMEM. And placed coverslips, imaging was performed in an observation chamber in DMEM without phenol red, supplemented with 25 mM HEPES. Following a 3-minute baseline acquisition, add the required bacteria strains and OD600 into the chamber. Following a 10-minute incubation at room temperature to allow bacterial attachment. Imaging was then carried out at 35°C to allow for bacterial type III secretion, using a Nikon Eclipse TE200 inverted fluorescence microscope with a 60× objective for 1h of infection. Fluorescence signals were acquired 485 nm with excitation and 535 nm emission parameters. Image control and data acquisition were managed by Simple32 software (Compix Inc.), depending on the experiment requirement, imaging at 22ms, 57ms or 10 sec acquisition intervals. 3 μM of Histamine and 2 μM of ionomycin were applied to check the ability of cells to show calcium response. Images were captured using a CMOS camera (Hamamatsu) and analyzed using the same software.

Immunofluorescence analysis

Cells were washed three times with PBS and fixed with 3.7% paraformaldehyde (PFA), permeabilized with 0.1% Triton X-100 for 5 minutes and washed with PBS. Blocking was performed using 3% FBS in PBS for 30 minutes at room temperature. Cells were incubated with anti-Z0-1 polyclonal antibody (ThermoFisher Scientific, # 40-2200) for 1 hour at a 1:50 dilution in PBS containing 1% FBS for an hour, followed by anti-rabbit IgG-Alexa 488 (Life Technologies, #A11034) or mouse phalloidin-Alexa 488 (Fischer Scientific, #17511176) at a 1:200 dilution and DAPI (1 μg/ml; Merck, #102362276001) for another hour. Samples were mounted using Dako mounting medium (Agilent) and imaged using a Nikon Ti2 confocal microscope equipped with a 60× objective and Nikon acquisition software.

Modelling

We develop a fully stochastic spatial model to simulate Ca2+ release dynamics from IP3R clusters in a two-dimensional representation of a HeLa cell. The simulation domain measures 10 × 10 µm2 and is discretized into a 20 × 20 grid of compartments (0.5 × 0.5 µm2 each), each representing a cytosolic subvolume of 10-16 L. Ca2+ exchange between the Endoplasmic Reticulum (ER) and the cytosol occurs through IP3R-mediated release, SERCA uptake, and ER Ca2+ leak, following the framework by Voorsluijs et al., 2019 and Ornelas-Guevara et al., 2023. Ca2+ diffusion is implemented stochastically as a kinetic process between adjacent compartments (Kraus et al., 1996), with a diffusion coefficient of 100 µm2/s to reflect moderate endogenous buffering. Each IP3R cluster functions as a single unit with four possible states: Open (O), Closed (C) and two Inhibited states (i1, and i2), transitioning in response to local [Ca2+] and [IP3]. This phenomenological description captures key characteristics of Ca2+ puffs and their transition to global signals via Ca2+ diffusion and Calcium Induced Calcium Release.

We perform all simulations using the Gillespie algorithm, where each event, reaction or diffusion, is selected stochastically based on its propensity. See Supplementary Information S1 for additional information about the model.

Western Blot analysis

To obtain total cell extracts, cells were lysed in sample buffer 1x (62.5 mM Tris pH=8, 2% SDS, 10% glycerol, 0.05% bromophenol blue, 5% β- mercaptoethanol) and boiled at 95°C for 5 minutes. Proteins from total lysates were separated by SDS PAGE and transferred to nitrocellulose membrane (0.45μM AmershamTM ProtranTM). Western Blot analysis was performed according to standard procedure using the following primary antibodies diluted in PBS containing 0.1 % Tween-20 and 5 % non-fat milk: IκB-alpha (OZYME, #9424S) at a 1:1000 dilution, Phospho-IκB-α (OZYME, #9424S) at a 1:1000 dilution, NF-κaB p65 (OZYME, #9246S) at a 1:5000 dilution, Phospho-NF-κB p65 (OZYME, #3033S) at a 1:1000 dilution, HSP90 (Santa Cruz Biotechnologies #sc-13119) at a 1:1000 dilution. The secondary HRP-conjugated anti-mouse (Cytiva) and anti-rabbit (Sigma) antibodies were used at a 10-4 dilution.

Immunoprecipitation assays

HeLa cells were seeded in 150 cm2 dishes (7.4 × 10⁶ cells/dish) and cultured in DMEM supplemented with 10% FBS. Cells were pretreated with 20 µM BAPTA for 30 minutes at room temperature where indicated, then stimulated with 10 ng/mL TNF-α alone or in combination with 150 nM ATP for an additional 12 minutes. After treatment, cells were washed with ice-cold PBS containing 1 mM NaF and 1 mM Na₃VO₄, lysed in ice-cold lysis buffer (50 mM Tris-HCl pH 7.5, 0.5% Triton X-100, 100 mM NaCl, 1 mM DTT, protease inhibitor cocktail without EDTA), and incubated on a rotating wheel at 4°C for 1 hour. Lysates were clarified by centrifugation at 13,000 rpm for 30 minutes at 4°C. Supernatants were incubated with 5 µL of anti-NF-kB p65 antibody (Abcam, ab16502) for 2 hours at 4°C with rotation, followed by overnight incubation with pre-equilibrated protein A/G beads. Immunocomplexes were washed once with lysis buffer and twice with PBS, then eluted in Laemmli buffer by boiling at 100°C for 5 minutes. Input and IP samples were analyzed by Western blotting using antibodies against O-GlcNAc (Abcam, ab2739), anti-NF-kB p65 (OZYME, 9246S), and Actin (OZYME, 4967).

Statistical Analysis

All quantitative data are presented as mean ± SEM from at least three independent experiments. Statistical significance was assessed using unpaired two-tailed Student’s t-tests with unequal variance, unless otherwise specified. GraphPad Prism 7 (GraphPad Software) was used for statistical analysis, and p-values < 0.05 were considered statistically significant.